\documentclass[a4paper, 12pt, oneside]{memoir}
\usepackage{morewrites}
\usepackage{fontspec}
\usepackage[T1]{fontenc}
%\usepackage[utf8]{inputenc}
\usepackage{lipsum}
\usepackage{titlesec}
\usepackage{titling}
\usepackage{geometry}
\usepackage[canonical, leipzig]{baarux}
\usepackage{vowel}
\usepackage{phonrule}
\usepackage{float}
\usepackage{chngcntr}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{verbatim}
\usepackage[graphicx]{realboxes}
\usepackage{tikz}
\usepackage{tikz-qtree}
\usetikzlibrary{patterns,positioning, decorations.pathmorphing, fit}
\usepackage{phonrule}
\usepackage[]{graphicx}
\usepackage{caption}
\usepackage[]{amsmath}
\usepackage{amssymb}
\usepackage{titletoc,tocloft}
\usepackage{chngcntr}
\usepackage{etoolbox}
\usepackage[block]{leipzig}
\usepackage{tabularx, ragged2e}
\usepackage{makecell}
\usepackage[hidelinks]{hyperref}
\usepackage{comment}
\usepackage[citestyle=authoryear-comp, backend=biber]{biblatex}
\addbibresource{refs.bib} 

\DeclareDelimFormat[textcite]{postnotedelim}{\addsemicolon\space}
\DeclareDelimFormat[parencite]{postnotedelim}{\addcolon\space}

\DeclareFieldFormat{postnote}{\mknormrange{#1}}
\DeclareFieldFormat{multipostnote}{\mknormrange{#1}}

%\DoubleSpacing
\linespread{1.75}

\usepackage{expex}
\lingset{Everyex={\SingleSpacing}}

\setlength\parindent{0pt}

\renewcommand{\arraystretch}{1}


\def\trstyle#1{%
  `\textit{#1}'%
}
\begin{comment}
\baaruset{%
  gloss.markup={},
  tr.markup=\trstyle
}
\end{comment}
\baaruset{source.markup=}
\newbaarulinetype{rightcomment}{manner}{\baaruparens}


\setmainfont{Charis SIL} 
\newcommand{\conlang}[1]{\textit{#1}}
\newcommand{\family}{Hapi-Guilené }
\newcommand{\peoplef}{Hapi }
\newcommand{\emh}[1]{\textit{#1}}

\newcommand{\fourcolumn}[4]{
        \begin{minipage}{0.225\textwidth}
                #1
        \end{minipage}%
        \hfill
        \begin{minipage}{0.225\textwidth}
                #2
        \end{minipage}
        \hfill
        \begin{minipage}{0.225\textwidth}
                #3
        \end{minipage}%
        \hfill
        \begin{minipage}{0.225\textwidth}
                #4
        \end{minipage}
}

\setcounter{secnumdepth}{4}

\titleformat{\chapter}[hang]%
    {\large\normalfont\bfseries\centering}% global formatting (number and title)
    {\thechapter}% label: number and its formatting
    {0.5em}% spacing between number and title
    {}% optional (content between number and title)
\titleformat{\section}[hang]%
    {\large\normalfont\bfseries}% global formatting (number and title)
    {\thesection}% label: number and its formatting
    {0.5em}% spacing between number and title
    {}% optional (content between number and title)
\titleformat{\subsection}[hang]%
    {\large\normalfont\bfseries}% global formatting (number and title)
    {\thesubsection}% label: number and its formatting
    {0.5em}% spacing between number and title
    {}% optional (content between number and title)
\titleformat{\subsubsection}[hang]%
    {\large\normalfont\bfseries}% global formatting (number and title)
    {\thesubsubsection}% label: number and its formatting
    {0.5em}% spacing between number and title
    {}% optional (content between number and title)
\titleformat{\paragraph}[hang]%
    {\large\normalfont\bfseries}% global formatting (number and title)
    {\theparagraph}% label: number and its formatting
    {0.5em}% spacing between number and title
    {}% optional (content between number and title)


\AtBeginDocument{\def\theexi{\arabic{exi}}}


\makeglossaries

\newleipzig{hab}{hab}{habitual aspect}
\newleipzig{perf}{perf}{perfective aspect}
\newleipzig{instr}{instr}{instrument}
\newleipzig{quotative}{quot}{quotative}
\newleipzig{redup}{r}{reduplication}  
\newleipzig{seq}{seq}{sequential }
\newleipzig{fourth}{4}{fourth person}
\newleipzig{nmz}{nmz}{nominalizer}
\newleipzig{c}{c}{complementizer}
\newleipzig{intrg}{intrg}{interrogative mode}
\newleipzig{ego}{egoph}{egophoric}
\newleipzig{dep}{dep}{dependent}
\newleipzig{vis}{vis}{visual}
\newleipzig{nvis}{nvis}{non-visual}
\newleipzig{infer}{infer}{inferred}
\newleipzig{evid}{evid}{evidential}
\newleipzig{agent}{A}{agent}
\newleipzig{subject}{S}{subject}
\newleipzig{patient}{P}{patient}
\newleipzig{rpasto}{rec.past}{recent past}
\newleipzig{rpastt}{int.past}{intermediate past}
\newleipzig{dpasto}{dist.past}{distant past}
\newleipzig{dpastt}{rem.past}{remote past}
\newleipzig{rpast}{past1}{primary past}
\newleipzig{dpast}{past2}{secondary past}
\newleipzig{prl}{perl}{perlative}
\newleipzig{pvt}{priv}{privative}
\newleipzig{smb}{sembl}{semblative}
\newleipzig{ont}{on.top}{'on top' case}
\newleipzig{bel}{below}{'below' case}
\newleipzig{next}{next.to}{'next to' case}
\newleipzig{tsf}{3sf}{third person singular female}
\newleipzig{tsm}{3sm}{third person singular male}
\newleipzig{tsn}{3sinan}{third person singular inanimate}
\newleipzig{tp}{3p}{third person plural}
\newleipzig{tsposs}{3spos}{third person singular possessive}
\newleipzig{ff}{1}{first person}
\newleipzig{fs}{1s}{first person singular}
\newleipzig{ss}{2s}{second person singular}
\newleipzig{ssap}{2}{second person}
\newleipzig{st}{2/3}{second/third person}
\newleipzig{tt}{3}{third person}
\newleipzig{ts}{3s}{third person singular}
\newleipzig{fat}{1+3}{first person plural exclusive}
\newleipzig{fas}{1+2}{first person plural inclusive}
\newleipzig{aug}{aug1}{augmentative 1}
\newleipzig{auga}{aug2}{augmentative 2}
\newleipzig{dim}{dim}{diminutive}
\newleipzig{fsposs}{1spos}{first person singular possessive}
\newleipzig{fpposs}{1ppos}{first person plural possessive}
\newleipzig{ssposs}{2spos}{second person singular possessive}
\newleipzig{ssh}{2hon}{second person singular honorific}
\newleipzig{sshposs}{2hpos}{second person singular honorific possessive}
\newleipzig{tpposs}{3ppos}{third person plural possessive}
\newleipzig{cl}{cl}{classifier}
\newleipzig{link}{link}{linking particle}
\newleipzig{proximal}{prox}{proximal deixis}
\newleipzig{distal}{dist}{distal deixis}
\newleipzig{hort}{hort}{hortative}
\newleipzig{opt}{opt}{optative}
\newleipzig{obligative}{obl}{obligative modal}
\newleipzig{qu}{q}{question marker}
\newleipzig{posit}{posit}{positive}
\newleipzig{negat}{negat}{negative}
\newleipzig{impv}{impv}{imperative}
\newleipzig{epist}{epist}{epistemic}
\newleipzig{juss}{juss}{jussive mode}
\newleipzig{recip}{recip}{reciprocal}
\newleipzig{cont}{cont}{continuative}
\newleipzig{itr}{itr}{intransitive}
\newleipzig{trans}{trans}{transitive}
\newleipzig{topic}{topic}{topic}
\newleipzig{noct}{noct}{nocturnal}
\newleipzig{diur}{diur}{diurnal}
\newleipzig{ds}{ds}{different subject switch-reference}
\newleipzig{punct}{punct}{punctual}
\newleipzig{frust}{frust}{frustrative}
\newleipzig{rep}{rep}{reportative}
\newleipzig{inch}{inch}{inchoative}
\newleipzig{crast}{crast}{crasternal}
\newleipzig{mir}{mir}{mirative}

\glsfindwidesttoplevelname[\leipzigtype]

\title{A Grammar of lang5}
\author{u/tryddle}
\date{May 2020}


\newcommand{\people}{Hapi }

\newlength\drop
\makeatletter
\newcommand*\titleM{\begingroup% Misericords, T&H p 153
\setlength\drop{0.08\textheight}
\centering
{\scshape Dedicated to G.D., J.K., P.A., T.F., C.L., and many others\par}
\vspace*{\drop}
   
\vspace*{\drop}
\begin{flushleft}
{\HUGE\bfseries Hapi}\\
\end{flushleft}
\vspace*{\drop}
\vspace*{\drop}
\begin{flushright}
{\LARGE a constructed language by u/tryddle}\\[\baselineskip]
{\scshape a reference grammar}\\[\baselineskip]
{\scshape \@date}\par
\end{flushright}


\endgroup}
\makeatother



\begin{document}
\begin{titlingpage}
\titleM{}
\end{titlingpage}
\pagebreak
\newgeometry{margin=2.5in}

\vspace*{\fill}

\begin{center}
    \begin{verbatim}
        Version: 0.1.7-alpha
        Date: 22 March, 2021
    \end{verbatim}
    ©opyright April 2020 tryddle
\end{center}

\vspace*{\fill}

\restoregeometry

\pagebreak
\setcounter{tocdepth}{4}
\tableofcontents
\newpage 
\listoffigures
\listoftables
\newpage
\printglossaries
\newpage

\chapter{Introduction}
This preliminary reference grammar is the product of the attempt at documenting the Hapi language, a constructed language made by me, u/tryddle. In my just over 3 ½ years of conlanging, I've learned much, but not only because of my own ability to insert information into my brain, but also because of many people that have supported and helped me on the way. Without them, and most probably without you, the interested reader, this whole language would probably not exist. When I first discovered conlanging as an art form in late 2017, I was interested — immediately. Thanks to Artifexian\footnote{https://www.youtube.com/user/Artifexian} and his youtube channel, as well as Mark Rosenfelder and his amazing book `The Language Construction Kit', I was introduced into the rabbit hole of linguistics, more specifically the rabbit hole of language construction. 
The goal of this document is not only to document the Hapi language in all of its complexity, but also to document a language only using the markdown language \LaTeX; a challenge which I've failed to achieve so many times before. 
But now onto the most important part of this introduction. Obviously I want to thank many people who have helped me on this journey. Primarily, I especially thank Gordon Daws, Jacob Kronenberg, Paul Daly, Tobias Fernandez and Carl Leon, to whom this reference grammar is dedicated. They have supported me in so many ways, not all concerning conlanging, that it would have been heresy to not include them in these acknowledgements. Special thanks to Akam Chijir, who has helped me in many ways while I was battling the uncanny complexity of \LaTeX.
\chapter{Background of the study}
The following study was conducted during my three-year stay in the Kanangan rainforest, more specifically in Hapi territory. During this time, I learned a lot about the Hapi language and its people. In this chapter, I shall consider the background of this study, explaining some cultural background, as well as some typological characteristics. In section \ref{snmc} I will discuss the nomenclature of the language's name, starting with the term `Hapi', and then moving onward to several terms of self-reference, as well as exonymic terms of reference. In section \ref{sorigins} I will present the origins and the historical distribution of the Hapi language, as well as give some insight into the \family language family. Then, in section \ref{sresearch} I will discuss some previous research that has been conducted on the language. In section \ref{stypo} I will consider a quick typological overview of the language, to give the reader an insight what is about to be discussed. After a quick language vitality assessment in section \ref{svitality}, I will give an overview of the present study, presenting its content in a compact manner.
The Hapi language is spoken by around 250 people in the Kanangan rainforest, near the Palhen tributary of the Kanang river. They live in huts called the \emh{koí} and each village consists of 20-40 people. Women and men live separately, the woman with their children and female teens and the men with the male teens. They worship multiple dieties or spirits, with most of them taking on the form of animals or geographical landmarks. Further ethnographical studies have to be conducted on the culture of the \peoplef people, and I encourage every reading ethnologist to help making progress in the documentation of these people. 

\section{Nomenclature}\label{snmc}
\subsection{The Term `Hapi'}
The term `Hapi' is only an exonymic term of reference given to the \peoplef people and their language by the Yamonari people, which derives from the term \emh{hap-i} in Yamarri, the language of the Yamonarï, which approximately can be translated to the meaning `needle people'. This is most probably a reference to the cosmetic needles and piercings the \peoplef often insert into their noses, lips and ears to depict prosperity and/or wealth. The Yamonari and the \peoplef people are two adjacent people which have interacted with each other several times, but generally live very separated and isolated, as they are the only tribes in the region of the Palhen river. 
The first appearance of the term `Hapi' was in the Ataman encyclopedia \textit{ndéke7undu}, where the ethnicity is described as ``a peaceful tribe living in the midst of the forest known as the forest of \textit{kámgá7}''.\footnote{A term coined by the Ataman geographers meaning `large forest'.} However, \emh{hapi} as a lexeme was later borrowed into the Hapi language with the meaning `people' or `person'. Therefore, while the term itself is originally exonymic, it became an endonym quite early on.
\subsection{Other terms of reference and ethnic diversity}
`Hapi' is however not the only term of self-reference among the Hapi people. There is also the term \emh{tàahopí}, which has its root in the first person exclusive pronoun of the same form; additionally, there is \emh{hòkaapasa}, a term meaning `hunters' or `hunting people'.
There are several other exonyms which were used in the past and are still being used as of today. The Kʰʉtɛɛ of the northern Kɨɨlmi tributary refer to the Hapi as \emh{kʰəkdʼəhway}, the `nonsense-speakers' and the Pañhanipe call them \emh{ñaginhap}, `those who live downstream'. Nevertheless there are also some  pejorative exonyms, such as the Ikoșh term \emh{șhasti}, simply meaning `enemy', or the Apéhigɨ exonym \emh{hapitéa}, which itself contains the term `Hapi' and means `meat-eater'.\footnote{This Apéhigɨ exonym has been absorbed into the language's vocabulary, retaining its old meaning. In recent times there have been some discussions about the racist nature of the term, and whether it should be erased from dictionaries.}
\section{Origins and classification}\label{sorigins}

\section{Previous research}\label{sresearch}
\section{Typological characteristics}\label{stypo}
\section{Language vitality assessment}\label{svitality}
\section{The present study}
\chapter{Phonology}
In this chapter, I will concentrate on the phonological aspects of Hapi, and present the main features of its phonology. In section \ref{scons}, I will focus on the language's consonants, section \ref{svow} discusses vowels, section \ref{sphond} gives the phonetic description of these phonemes and section \ref{sall} focuses on the allophony of these sounds. After that, we will discuss the non-segmental phonology of Hapi, starting with the stress assignment system in section \ref{sstress}, then in section \ref{ssyll}, we will discuss the phonotactics and syllable structure, and finally concluding with the tonal system of Hapi.  Throughout this chapter, examples are given, first in phonological transcription, and then in phonetic and orthographic transcriptions. Examples are presented in both the International Phonetic Alphabet (IPA) and the tentative Hapi orthographic system.
\section{Segmental Phonology}
In this section, I will present the language's phonemes, including its consonants and vowels, and their phonotactic distribution within the syllable. Furthermore I will give a list of minimal pairs for both consonants and vowels.
\subsection{Consonants}\label{scons}
In this section, I will present the language's consonant phonemes. There are 6 consonant phonemes in the Hapi language. Table \ref{cphon} presents the consonant sounds, while table \ref{cortho} depicts those sounds in the tentative Hapi orthography.
\begin{table}[H]
\centering
\begin{tabular}{@{}llll@{}}
\toprule
 & Peripheral & Alveolar & Non-Alveolar \\ \midrule
Stop & p$\sim$b & t$\sim$r & k$\sim$g \\
Fricative & h & s$\sim$ts & ʃ$\sim$ʂ$\sim$χ \\ \bottomrule
\end{tabular}
\caption{Consonant phonemes}
\label{cphon}
\end{table}
\begin{table}[H]
\centering
\begin{tabular}{@{}llll@{}}
\toprule
 & Peripheral & Alveolar & Non-Alveolar \\ \midrule
Stop & p & t & k \\
Fricative & h & s & x \\ \bottomrule
\end{tabular}
\caption{Tentative consonant orthography}
\label{cortho}
\end{table}
\subsection{Vowels}\label{svow}
In this section I will showcase the Hapi language's vowels. There are three phonemic vowels, which is significantly less than of most adjacent languages. Table \ref{vphon} gives an overview of these sounds, while table \ref{vortho} presents the tentative Hapi orthography of these sounds. Each vowel may also take one of three tones; this will be discussed in section \ref{stone}
\begin{table}[H]
\centering
\begin{tabular}{@{}lll@{}}
\toprule
 & Front & Back \\ \midrule
High & i & \multirow{2}{*}{o} \\
Low & a &  \\ \bottomrule
\end{tabular}
\caption{Vowel phonemes}
\label{vphon}
\end{table}
\begin{table}[H]
\centering
\begin{tabular}{@{}lll@{}}
\toprule
 & Front & Back \\ \midrule
High & i & \multirow{2}{*}{o} \\
Low & a &  \\ \bottomrule
\end{tabular}
\caption{Tentative vowel orthography}
\label{vortho}
\end{table}
\subsection{Phonetic description of phonemes}\label{sphond}
\subsubsection{Consonants}
In this section I will consider the phonetic description of consonants in Hapi. Allophonic variations are discussed in section \ref{sall}. All consonants may appear in word-intial and in word-medial position. Since all syllables in the language are open\footnote{Note that the onset may be filled by /h/, in which case either the subsequent segment is preaspirated, or it is debuccalized to /ʔ/. Thus, most if not all syllables can be analyzed as being open.}, there is no restriction on where a consonant segment may be placed. In the following section, examples for each position and each consonant will be given.
\newline
/p$\sim$b/ is a voiceless or voiced bilabial stop. It is pronounced as /b/ by most children and women, while it is pronounced /p/ by most men and the elderly. It appears in both word-initial and word-medial position. Henceforth it will be standardized to /p/.

\begin{columns}
\cols /pàah/ & [pʰaːʔ˩˧] & \emh{pàah} & `small river'
\cols /ʂopàì/ & [ʂo˧pai̯˩] & \emh{xopàì} & `rotten wood'
\end{columns}



/t$\sim$r/ is a voiceless alveolar stop or a voiced alveolar trill. It is pronounced as /r/ by most children and women, while it is pronounced /t/ by most men and the elderly. It appears in both word-initial and word-medial position. Henceforth it will be standardized to /t/.

\begin{columns}
    \cols /táosiiʂó/ & [tʰao̯˥˧ʃiː˧ho˥] & \emh{táosiihó} & `male cousin'
    \cols /hótahóoi/ & [ho˥ta˧oː˥˧i˧] & \emh{hótahóoi} & `black iguana'
\end{columns}


/k$\sim$g/ is a voiceless or voiced velar stop. It is pronounced as /g/ by most children and women, while it is pronounced /k/ by most men and the elderly. It appears in both word-initial and word-medial position.  Henceforth it will be standardized to /k/.

\begin{columns}
    \cols // & [] & \emh{} & `'
    \cols // & [] & \emh{} & `'
\end{columns}


/h/ is a voiceless glottal fricative. It appears in word-initial and  word-medial position. 

\begin{columns}
    \cols // & [] & \emh{} & `'
    \cols // & [] & \emh{} & `'
\end{columns}

/s$\sim$ts/ is a voiceless alveolar fricative or affricate. It is pronounced as /s/ by most children and women, while it is pronounced /ts/ by most men and the elderly, especially in emphasized speech. 
It appears in both word-intial and word-medial position. Henceforth it will be standardized to /s/.

\begin{columns}
    \cols // & [] & \emh{} & `'
    \cols // & [] & \emh{} & `'
\end{columns}

/ʃ$\sim$ʂ$\sim$χ/ is a voiceless alveolar or retroflex or uvular fricative. It is pronounced as /ʃ/ by most children, as /ʂ/ by most women and the elderly, and as /χ/ by men and in emphasized speech. It is sometimes debuccalized to /h/ in rapid speech. It appears in world-initial and word-medial position. Henceforth, it will be standardized to /ʂ/

\begin{columns}
    \cols // & [] & \emh{} & `'
    \cols // & [] & \emh{} & `'
\end{columns}

\subsubsection{Vowels}
In this section I will consider the phonetic description of consonants in Hapi. Vowels may appear in the nucleus of a syllable as short vowel, as long vowel or as a glide.\footnote{This excludes /a/, since it never appears as a glide. Compare \newline \fourcolumn{/hoa/}{[ho̯a]}{\emh{hoa}}{3\textsc{sg}} \newline \fourcolumn{/siikào/}{[siːkao̯]}{\emh{siikào}}{`canoe'} \newline This may be related to a diachronic approach where /o/ might have been /u/ in the past, changing its realizations slowly over the course of many years. If that was the case, a glide pair like /j w/ would have been very reasonable.} In the following section, examples for each appearance of each vowel will be given.

/i/ is a high front unrounded vowel. It may appear as short or long vowel, or as a glide.

\begin{columns}
    \cols // & [] & \emh{} & `'
    \cols // & [] & \emh{} & `'
    \cols // & [] & \emh{} & `'
\end{columns}

/o/ is a close-mid back rounded vowel. It may appear as short or long vowel, or as a glide.

\begin{columns}
    \cols // & [] & \emh{} & `'
    \cols // & [] & \emh{} & `'
    \cols // & [] & \emh{} & `'
\end{columns}

/a/ is a low front unrounded vowel. It may appear as short or long vowel.

\begin{columns}
    \cols // & [] & \emh{} & `'
    \cols // & [] & \emh{} & `'
\end{columns}

\subsection{List of Minimal Pairs}\label{smin}
In the following section I will give a list of minimal pairs for both consonants and vowels.
\subsubsection{Consonants}

\begin{columns}
    \cols \emh{háiki} & `type of nut' & \emh{haípi} & `soup'
    \cols \emh{kaxá} & `to eat' & \emh{saxá} & \textsc{cl}:palm.tree.trunk
    \cols \emh{paáti} & `bird' & \emh{pahi} & `bird'
\end{columns}

\subsubsection{Vowels}
Firstly I will consider minimal pairs of vowel quality, then I will move onto vowel length minimal pairs and finally I will present tonal minimal pairs.

\begin{columns}
    \cols \emh{hì} & \textsc{cl}:time & \emh{hó} & `man's name'
    \cols \emh{kahoa} & `to build' & \emh{kahoó} & `big tree'
    \cols  \emh{sáhaa} & `to be grateful' & \emh{sahóó} & `to carve'
\end{columns}

\begin{columns}
    \cols \emh{hì} & \textsc{cl}:time & \emh{hó} & `man's name'
    \cols \emh{kahoa} & `to build' & \emh{kahoó} & `big tree'
    \cols  \emh{sáhaa} & `to be grateful' & \emh{sahóó} & `to carve'
\end{columns}

\begin{columns}
    \cols \emh{hakó} & `ago' & \emh{hákoo} & `\textsc{obl}' 
    \cols \emh{hása} & `loincloth' & \emh{hasáa} & `to be the first'
    \cols  \emh{káhí} & `turtle' & \emh{kahíí} & `boat'
\end{columns}

\begin{columns}
    \cols \emh{hoi} & \textsc{cop:loc} & \emh{hói} & `to give' 
    \cols \emh{hóika} & `to build' & \emh{hóíka} & `big tree'
    \cols  \emh{ò} & `and' & \emh{ó} & \textsc{seq}
\end{columns}

\subsection{Allophonic Variations}\label{sall}
The following section discusses all phonological processes that alter the phonetic realizations of phonemes in a remarkable way. This does not include morphophonological changes. 
\subsubsection{Stop Allophony}
\textbf{1)} /p/ is realized as [ç] before the high vowel /i/.
\begin{examples}
\ex \phonr{p}{ç}{i}
\end{examples}
\textbf{2)} /p/ is aspirated at the beginning of a word.
\begin{examples}
\ex \phonl{p}{pʰ}{\#}
\end{examples}
\textbf{3)} /b/ is realized as [ʝ] or [ɟ] before /i/.
\begin{examples}
\ex \phonr{b}{ʝ, ɟ}{i}
\end{examples}
\textbf{4)} /b/ is devoiced at the beginning of a word.
\begin{examples}
\ex \phonl{b}{p}{\#}
\end{examples}
\textbf{5)} /t/ is aspirated at the beginning of a word.
\begin{examples}
\ex \phonl{t}{tʰ}{\#}
\end{examples}
\textbf{6)} /r/ is devoiced at the beginning of a word.
\begin{examples}
\ex \phonl{r}{r̥}{\#}
\end{examples}
\textbf{7)} /k/ is strengthened to [kx] at the beginning of a word and in a stressed syllable.
\begin{examples}
\ex \phonc{k}{kx}{
    \oneof{
    \#\phold\\
    \phonfeat{
        +stress
    }
    }
}
\end{examples}
\subsubsection{Fricative Allophony}
\textbf{8)} /h/ is realized as [ç] before /i/.
\begin{examples}
\ex \phonr{h}{ç}{i}
\end{examples}
\textbf{9)} /h/ is elided between two distinct vowels.
\begin{examples}
\ex \phonb{h}{∅}{V₁}{V₂}
\end{examples}
\textbf{10)} /h/ is realized as [ʔ] at the end of a word.
\begin{examples}
\ex \phonc{h}{ʔ}{
    \oneof{
    \phold \# \\
    \textit{[h] elsewhere}}
    }
\end{examples}
\textbf{11)} /s/ is palatalized to [ʃ] before /i/.
\begin{examples}
\ex \phonc{s}{ʃ}{
    \oneof{
    \phold i \\
    \textit{[s] elsewhere}}
    }
\end{examples}
\textbf{12)} /ʂ/ is debuccalized to /h/ in between two distinct vowels.
\begin{examples}
\ex \phonc{ʂ}{h}{
    \oneof{
    V₁ \phold V₂ \\
    \textit{[ʃ$\sim$ʂ$\sim$χ] elsewhere}}
    }
\end{examples}
\subsection{Conclusion}
In what has preceded, I have explained the segmental phonology of Hapi, starting with the consonant and vowel phonemes in section \ref{scons} and \ref{svow}. Then I posited a phonetic description of these segments in \ref{sphond}, while presenting a list of minimal pairs in section \ref{smin}. Finally, I have considered the allophonical processes that shape the phonetic realizations of phonological words. Thus, in the following sub-sections, stress, as well as syllable structure and tone, all three suprasegmental features of Hapi, will be discussed. 
\section{Stress}\label{sstress}
Stress in Hapi is very predictable, since it invariably falls on the penultimate syllable of the stem. As can be seen in examples (\ref{stress1}) and (\ref{stress2}), conjunct suffixation does not change the stress distribution, still yielding an expected stress pattern.

\begin{columns}\label{stress1}
    \cols /pàa.ʂo/ & [ˈpa˩˧.ho˧] & \emh{pàaxo} & `tendon'
    \cols /pàa.ʂo.áh/ & [ˈpaː˩˧.ho˧.aʔ˥] & \emh{pàaxoáh} & `a large tendon'
\end{columns}

\begin{columns}\label{stress2}
    \cols /hó.sii/ & [ˈho˥.ʃiː˧] & \emh{hósii} & `son' 
    \cols /hó.sii.ʂii/ & [ˈho˥.ʃiː˧.hiː˧] & \emh{hósiixii} & `grandson' 
\end{columns}



\section{Syllable Structure and Phonotactics}\label{ssyll}
In the following section I will consider the syllable structure of the Hapi language.
The language accepts open syllables of the types V, CV, CVh, CVV and CVVh. Phonetically closed syllables are not attested. The syllable structure of Hapi is presented in (\ref{fig:syllstruc}), where $\tau$ stands for the contour tone of the syllable.

\begin{figure}[H]
    \centering
        \begin{tikzpicture}
\Tree [.$\sigma$ 
          [\node[style={align=center}]{(C)};]
          [\node[style={align=center}]{V};]
          [\node[style={align=center}]{(V)};]
          [\node[style={align=center}]{(h)}; ]
          \edge[dashed];
          [\node[style={align=center}](tone){$\tau_1$,($\tau_2$)};]
        ]
\end{tikzpicture}
    \caption{Syllable Structure}
    \label{fig:syllstruc}
\end{figure}
\subsection{Types of syllable structure}
In the following section I will give examples for each type of syllable in the Hapi language, of which there are five different types, namely: V, CV, CVh, CVV and CVVh. 
Firstly, the simplest of all possible syllables is V. (\ref{Vex}) gives an example for this syllable type. 

\begin{columns}\label{Vex}
    \cols /à/ & [a˩] & \emh{à} & \textsc{link}
\end{columns}


This grammatical particle is depicted formally in figure (\ref{fig:Vex}).

\begin{figure}[H]
    \centering
    \begin{tikzpicture}[sibling distance=60pt]
        \Tree [.$\sigma$ 
                [.V
                \node[style={align=center}]{a};
                ]
               \edge[dashed]; [\node[style={align=center}]{$\tau=$L};]
                ]
    \end{tikzpicture}
    \caption{Syllable Structure: Example 1}
    \label{fig:Vex}
\end{figure}

Then, filling in the onset yields the next syllable type, CV. Example (\ref{CVex}) is an example for this syllable type.

\begin{columns}\label{CVex}
    \cols /kó/ & [ko˥] & \emh{kó} & 3\textsc{sm}
\end{columns}

The phonological form of this pronoun is depicted formally in figure \ref{fig:CVex}

\begin{figure}[H]
    \centering
    \begin{tikzpicture}[sibling distance=40pt]
        \Tree [.$\sigma$ 
                [.C
                \node[style={align=center}]{k};
                ]
                [.V
                \node[style={align=center}]{o};
                ]
               \edge[dashed]; [\node[style={align=center}]{$\tau=$H};]
                ]
    \end{tikzpicture}
    \caption{Syllable Structure: Example 2}
    \label{fig:CVex}
\end{figure}

We can now add another vowel to the nucleus, yielding a syllable of the type CVV and a new tone slot, as depicted in example (\ref{CVVex}) and figure \ref{fig:CVVex}.\footnote{Note that syllables of the type VV are also possible. The author has decided not to list this rarely occuring feature.}

\begin{columns}\label{CVVex}
    \cols /pói/ & [poi̯˥˧] & \emh{pói} & 2\textsc{s}
\end{columns}


\begin{figure}[H]
    \centering
    \begin{tikzpicture}[sibling distance=30pt]
        \Tree [.$\sigma$ 
                [.C
                \node[style={align=center}]{p};
                ]
                [.V
                \node[style={align=center}]{o};
                ]
                [.V
                \node[style={align=center}]{i};
                ]
               \edge[dashed]; [\node[style={align=center}]{$\tau=_1$H, $\tau_2=$M};]
                ]
    \end{tikzpicture}
    \caption{Syllable Structure: Example 3}
    \label{fig:CVVex}
\end{figure}

Now onto the last two syllable types: by filling in the coda with /h/, the syllable becomes closed, as can be seen in examples (\ref{CVhex}) and (\ref{CVVhex}) and in figures \ref{fig:CVhex} and 
\ref{fig:CVVhex}.

\begin{columns}\label{CVhex}
    \cols /sóh/ & [soh˥] & \emh{sóh} & `and, with'
\end{columns}

\begin{columns}\label{CVVhex}
    \cols /ʂòih/ & [ʂoi̯ʔ˩˧] & \emh{xòih} & `(my) brother'
\end{columns}




\begin{figure}[H]
    \centering
    \begin{tikzpicture}[sibling distance=30pt]
        \Tree [.$\sigma$ 
                [.C
                \node[style={align=center}]{s};
                ]
                [.V
                \node[style={align=center}]{o};
                ]
                [.h
                \node[style={align=center}]{h};
                ]
               \edge[dashed]; [\node[style={align=center}]{$\tau_1=$H, $\tau_2$=M};]
                ]
    \end{tikzpicture}
    \caption{Syllable Structure: Example 4}
    \label{fig:CVhex}
\end{figure}

\begin{figure}[H]
    \centering
    \begin{tikzpicture}[sibling distance=20pt]
        \Tree [.$\sigma$ 
                [.C
                \node[style={align=center}]{x};
                ]
                [.V
                \node[style={align=center}]{o};
                ]
                [.V
                \node[style={align=center}]{i};
                ]
                [.h
                \node[style={align=center}]{h};
                ]
               \edge[dashed]; [\node[style={align=center}]{$\tau_1=$L, $\tau_2$=M};]
                ]
    \end{tikzpicture}
    \caption{Syllable Structure: Example 5}
    \label{fig:CVVhex}
\end{figure}

\section{Tone}\label{stone}
In this section I will discuss the tonal system of Hapi. The language distinguishes three tones, the high tone, the neutral middle tone and the low tone. All tones may occur in any combination within the syllable and the phonological word. The tentative Hapi orthography marks tone using diacritics: the high tone is represented by the acute ◌́, the low tone is depicted by the grave diacritic ◌̀ and the middle tone is left unmarked.
Examples (\ref{tone1} - \ref{tone3}) present instances of each tone.

\begin{columns}\label{tone1}
    \cols /káiʂo/ & [kxai̯˥˧ʂo˧] & \emh{káixo} & `(my) mother'
\end{columns}

\begin{columns}
    \cols /ʂapati/ & [ʂa˧pa˧ti˧] & \emh{xapati} & `string (of a bow)'
\end{columns}

\begin{columns}\label{tone3}
    \cols /tàhòhì/ & [tʰa˩ho˩çi˩] & \emh{tàhòhì} & `poison (for a dart)'
\end{columns}

Tones may also be realized as floating tones on certain morphemes, e.g. the 'on top' suffix \emh{-(V̀)xa}. In these cases, the preceding vowel may take the floating tone, or the preceding tone is overwritten, as can be seen in example (\ref{tone4})


\begin{columns}\label{tone4}
    \cols /táíʂo/ & [tai̯˥ʂo˧] & \emh{táíxo} & `stave'
    \cols /táíʂòʂa/ & [tai̯˥ʂo˩ʂa˧] & \emh{táíxòxa} & `stave (on top)'
\end{columns}


\section{Morphophonology}\label{smorphophono}
In this section I will consider various morphophonological processes that occur in the Hapi language. To start, I will clarify the terms of some morphophonological units such as the conjunct affix, the disjunct affix and the clitic in section \ref{sunitdefinition}. I will present some allomorphic variations in section \ref{sallomorphicvariations} that apply to all morphemes with the required environment, then moving onward to discuss each morpheme that exhibits notable morphophonological changes towards its root in section \ref{morphophonomorphemespecific}. Doing that, I will begin with the plural marker \Pl, then moving on to the diminutive marker {\Dim} in the sections \ref{morphophonoplural} and \ref{morphophonodim}. Afterwards, I will consider the morphophonology of the declarative disjunct suffix {\Decl} in section \ref{sdeclassim}.
\subsection{Definitions of morphophonological Units}\label{sunitdefinition}
In the following section I will examine the definitions of the morphophonological units employed in the Hapi language,  namely, the conjunct affix, the disjunct affix, the clitic and the particle. 
\subsubsection{Conjunct Affixes}
In the Hapi language, those affixes are classified as conjunct affixes which do constitute a single phonological word unit with the stem and attach on stem-level. However, they do not change the stress distribution within the word, so that stress invariably falls on a syllable that belongs to the stem. In interlinear gloss, this type of morphemic juncture is denoted by `-', as can be inferred by example (\ref{ex:conjunctaffix}).
\begin{examples}
\ex \label{ex:conjunctaffix}
\script ˈtʰiː˧ʃi˥aʔ˥
\bits tiisí -áh
\gloss amulet AUG
\tr the/a big (i.e. mighty) amulet
\end{examples}
\subsubsection{Disjunct Affixes}
While conjunct affixes attach to the stem and form a tight unit with it, disjunct affixes attach to the word unit on the word-level. However they're still more tightly bound to the phonological word as other words, as section \ref{smorphonodistinc} shows.  They do not change stress, and do not form a morphophonological unit with the word, so that rules like for example the debuccalization of /h/ at the end of a word apply to the affix-stem unit. In interlinear gloss, this type of juncture is marked by `=', as can be seen in example (\ref{ex:disjunctaffix}).\footnote{Note, that cross-linguistically, this sign is used to mark clitics; in this grammar however, `\texttt{==}' will be used to denote such phrase-level morphemes.}
\begin{examples}
\ex \label{ex:disjunctaffix}
\script ha˧ˈtaː˧˥çiʔ˧ko̯a˥˧
\bits ha- taápi -h =kóa
\gloss ANTIP paint FF DECL
\tr (I) am painting (something)
\end{examples}
\subsubsection{Clitics}
In Hapi grammar, clitics are morphemes which attach at phrase-level, i.e. cannot change stress but form a morphophonological unit with the word they attach to, whether it be the head or the dependent of a phrase. The clitic is marked by `\texttt{==}', and its usage is showcased in example (\ref{ex:clitic}).
\begin{examples}
\newbaarucmd{\cl}{\baarujuncture{\texttt{==}}}
\ex \label{ex:clitic}
\script ... ˈʃiː˥˧ti˧to˧
\bits tàhòhí -h kàah -xí =kóa síiti -∅ \cl to
\gloss poison ERG kill RPASTO DECL uakari ABS VIS
\tr (Earlier today,) the poison (of a dart) killed an uakari monkey, I saw that
\end{examples}
\subsubsection{Distinction from Words and ʔ-Insertion}\label{smorphonodistinc}
While the distinction word-clitic, or especially the difference between word and disjunct might appear shallow at first, upon further introspection into the phonological processes that appear on word boundaries, the difference might become clearer.
The main process that plays a significant role in this is the so-called ʔ-Insertion on empty onsets. This is a process that exclusively applies on word-initial empty onsets, which are then filled with a glottal stop [ʔ]. In example (\ref{ex:gsinsert}), an example is given for \textbf{a)} glottal stop insertion, and \textbf{b)} a proof on how disjuncts are distinct from conjuncts, more specifically, how certain allophonic processes only occur within the phonological word. For the sake of simplicity, I will not transcribe tone in this example.
\begin{examples}
\newbaarulinetype{aligned}{ipa}{}
\newbaarulinetype{aligned}{ans}{}
\newbaarucmd[ans]{=}{\baarujuncture{\texttt{=}}}
    \ex \label{ex:gsinsert}
    \words tàhopoí tóti xàahaó\MC2 kaihatoo póóihaó\MC3 páikòih\MC3 kóhsa\MC2 ítàahaká\MC4 hìhohihákoo\MC2
    \ans S Adv V Disj Adv [[V Aff]ₚ Disj]ₛ [Aff N Aff]ₚ N ∅ [[Aff V Aff]ₚ Disj]ₛ [Disj [V]ₚ]ₛ 
    \ipa tʰa.ˈo.poi̯ ˈtʰo.ti χaː hao̯ kxai̯.ˈa.toː pʰoː.i ʔ hao̯ pʰai̯ koi̯ ʔ kxoh.sa {∅} ʔi taː ha ka çi.o.çi ˈso.to
    \bits tàhopoí tóti xàa =haó kaihatoo póói -h =haó pái- kòi -h kóhsa -∅ í- tàa -a =ká hìhohi= hákoo
    \gloss FAS:AGENT ready COP \textsc{\textup{s/a>a(se)}} now arrive FF \textsc{\textup{s/a>a(se)}} SSHPOSS house DAT matter ABS INTRG AUX FF DECL delay OBLIGATIVE
    \tr Being ready, arriving at your house, should we (still) delay the affair?\footnote{In this example []ₚ indicates a phonological word, while []ₛ denotes the disjunct-word unit.} 
    %\source \parencite[75]{smoyd1309}
\end{examples}
Processes like k→kx, t→tʰ and p→pʰ, which are widely spread throughout the phonological realm, appear only near word-boundaries (namely, on the beginning of a word), and not on disjunct boundaries. An example is given in (\ref{ex:morphodistinc1}) and (\ref{ex:morphodistinc2}).
\begin{examples}
    \newbaarulinetype{aligned}{ipa}{}
    \ex \label{ex:morphodistinc1}
    \script so.ˈsaː.i kxo
    \ipa so.ˈsaː.i kxo
    \bits sosáai kó
    \gloss cat TSM
    \tr tomcat
    \ex \label{ex:morphodistinc2}
    \script çi.ˈki.ʃiʔ.ko̯a
    \ipa çi ˈki.ʃiʔ ∅ ko̯a
    \bits hi- kísih -h =kóa
    \gloss ANTIP dislike FF DECL
    \tr I dislike (it)
\end{examples}
\subsection{Allomorphic Variations}\label{sallomorphicvariations}
In this section I will discuss the basic parameters and morphophonological processes that apply to most, if not all bound morphemes in the Hapi language. The basic constraints or parameters of all bound morphemes are as follows. 
\subsubsection{Degemination of |h-h|}
Every context |h-h| is degeminated to |h|. This context is formalized in (\ref{ph:allom1}) and exemplified in (\ref{ex:ph:allom1}).
\begin{examples}
\ex \label{ph:allom1} \phon{h-h}{h}
\end{examples}
\begin{examples}
\ex\label{ex:ph:allom1}
\words sááí\MC2 háikia\MC2 kaai tà hóahkóa\MC3
\bits sáá =í háiki -a kaai tà hóah -h =kóa
\gloss eat \textsc{\textup{s/a>s(se)}} type.of.nut PL foul FS:SUBJECT vomit FF DECL
\tr After having eaten (some) foul nuts, I am (now) vomiting
\end{examples}
\subsubsection{Coalescence of morphemes in rapid speech}
Certain morphemes are fused in rapid speech, however there haven't been made any observations about these coalescence patterns. An example is given in (\ref{ex:rapidspeech}).
\begin{examples}
\ex \label{ex:rapidspeech}
\manner rapid speech
\words kaóxììhaíh\MC2 kíih hóíaka\MC3 xáháìh\MC2 xíoki
\bits kaóxììha -íh kíih hóí -a =ka xáháìi -h xíooki
\gloss teacher ERG old give ST DECL student DAT gift
\tr The old teacher gives a gift to a student
\end{examples}
\subsubsection{/h/-Insertion}
In each context of the form |VV-V|, /h/ is inserted on the morpheme boundary. This process is schematized in (\ref{ph:hinsert}) and exemplified in (\ref{ex:ph:hinsert}). In the context |V-VV| the same rule applies.
\begin{examples}
\ex \label{ph:hinsert} \phon{VV-V}{VVhV}
\ex \label{ex:ph:hinsert}
This section is work-in-progress
\end{examples}
\subsection{Morpheme-specific Processes}\label{morphophonomorphemespecific}
\subsubsection{Debuccalization of the marker \texorpdfstring{\Pl}{PL}}\label{morphophonoplural}
The debuccalization of the abbreviated version of the plural marker \Pl, namely the morpheme \emh{-s} is one of the most widely appearing processes in the Hapi language. The rule can be formalized as presented in (\ref{ph:plurals}). An instance of this process is exemplified in (\ref{ex:ph:plurals}).
\begin{examples}
\ex \label{ph:plurals} \phonr{s}{h}{C}
\ex \label{ex:ph:plurals}
\words ókoíhtah\MC4
\bits ó- koí -s -tah
\gloss SSPOSS hut PL PRL
\tr (we shall go) via your huts
\end{examples}
\subsubsection{The Diminutive marker \texorpdfstring{\Dim}{DIM}}\label{morphophonodim}
The diminutive marker \textsc{dim} undergoes two major processes: firstly, the deletion of the morpheme's onset, secondly, the dissimilation of the morpheme's nucleus. The deletion of the morpheme's onset, namely, of |ʂ|, appears in the context of a preceding /h/; this rule is formalized in (\ref{ph:dim}), an example is given in (\ref{ex:dim1}), but shall be reiterated here. 
\begin{examples}
\ex \label{ph:dim} \phonl{ʂ}{∅}{h}
\ex \label{ex:ph:dim}
\words hápaahii\MC2
\bits hápaah -xii
\gloss dog DIM
\tr the little dog
\end{examples}
The dissimilation of the morpheme's nucleus happens in the context of an /i/ in the preceding syllable's nucleus. This rule is depicted in (\ref{ph:dim2}) and exemplified in (\ref{ex:ph:dim2}).
\begin{examples}
\ex \label{ph:dim2} \phonl{iː}{o}{i(h).C}
\ex \label{ex:ph:dim2} 
\words hósiixo\MC2
\bits hósii -xii
\gloss son DIM 
\tr the/a grandson
\end{examples}
\subsubsection{Assimilation of the declarative affix \texorpdfstring{\Decl}{DECL}}\label{sdeclassim}
The declarative disjunct affix \Decl, \emh{=kóa}, is often reduced to \emh{=ká} or \emh{=ka} after a syllable containing /a/. This context is formalized in (\ref{ph:decl}) and an example is given in (\ref{ex:ph:decl}).
\begin{examples}
\ex \label{ph:decl} \phonl{kóa}{ka,ká}{a}
\ex \label{ex:ph:decl}
\words kó hásaaka\MC3 taahsáahi\MC2
\bits kó hásaa -a =kóa taah -sáahi
\gloss TSM:SUBJECT be.the.first ST DECL effort PVT
\tr He won (the race) without any effort
\end{examples}
\chapter{The Noun and the Noun Phrase}
This chapter discusses the noun and the noun phrase in the Hapi language. In the first section of this chapter, I will present the nominal morphology of the language. I will start by considering the nominal root and the noun structure in section \ref{snomroot}, then discussing gender marking in \ref{sgender}. I will move on and present the number marking, as well as diminutives and augmentatives in the sections \ref{snumber}, \ref{sdim} and \ref{saug}. In section \ref{spossession} I will consider the language's possessive system, first covering possessable, and then unpossessable nouns. In section \ref{scase} I will discuss the Hapi case system in all its complexity.
Moving onward onto section \ref{snounphrase} I will consider the structure of the noun phrase and its components, starting with the order of the noun phrase elements in \ref{snounorder} and an overview of the Hapi pronoun system in section \ref{spronouns}. After a comprehensive consideration of the language's classifier system in \ref{sclassifier}, I will lay out the different parts of the noun phrase in the sections \ref{sdemonstrative}, \ref{squantifier} and \ref{sattributes}.

\section{Nominal Morphology}
The internal structure of Hapi nouns is examined in this section, focussing on the grammatical categories encoded by nominal morphology i.e. gender, number, diminutives, augmentatives, possession and case.
\subsection{Nominal Root}\label{snomroot}
The structural properties of the nominal root include a set of suffixes responsible for the expression of number, diminutives, augmentatives, and case, while a set of prefixes is used to express possession on possessable nouns. Nominal roots can be divided into two morphophonological subtypes, which take different prefixes depending on the first segment in the root. If \textbf{(i)} the root starts with a vowel, then a certain set of prefixes is used, and if \textbf{(ii)} the root starts with a consonantal sound, another set is employed. 
\subsubsection{Internal structure of nominal roots}
% 17 mono, 67 disyllabic, 26 trisyllabic, 3 polysyllabic → 113 roots

In the following section I will consider the internal phonological structure of nominal roots, starting with the rare monosyllabic roots, then moving onward to the comparatively often found disyllabic, trisyllabic and polysyllabic, i.e. roots with more than four syllables, roots.
\paragraph{Monosyllabic roots}
There is a set of \textit{17} monosyllabic roots in the Hapi language, which are all considered to be very basic roots. There have been no reports of roots of the structure V and CV. Some examples for monosyllabic roots of the shape CV, CVh, CVV and CVVh are given in example (\ref{nomroot1ex}), (\ref{nomroot2ex}) and (\ref{nomroot3ex}).

\begin{columns}\label{nomroot1ex}
    \cols  /ʂòih/ & [ʂoi̯ʔ˩˧] & \emh{xòih} & `(my) brother'
\end{columns}
\begin{columns}\label{nomroot2ex}
    \cols  /píː/ & [çiː˥] & \emh{píí} & `major river'
\end{columns}
\begin{columns}\label{nomroot3ex}
    \cols  /kóih/ & [kxoi̯ʔ˥˧] & \emh{kóih} & `hideout'
\end{columns}


\paragraph{Disyllabic roots}
The majority of nominal roots in the Hapi language are disyllabic. In the analyzed corpus of \textit{113} nominal roots, there were \textit{67} disyllabic roots. 
\paragraph{Trisyllabic and polysyllabic roots}
There is a decent amount of tri- and polsyllabic nominal roots in the Hapi language. Trisyllabic roots are mostly of the shape CV.CV.CV, CV.CV.CVV, CV.CV.VV and CV.CVV.CV. Some examples for tri- and polysyllabic nominal roots are given in example (\ref{nomroot4ex}) - (\ref{nomroot6ex}).

\begin{columns}\label{nomroot4ex}
    \cols /hòkaapa/ & [ho˩kxaː˧pa˧] & \emh{hòkaapa} & `hunter'
\end{columns}
\begin{columns}\label{nomroot5ex}
    \cols /papáʂi/ & [pʰa˧pa˥hi˧] & \emh{papáxi} & `door'
\end{columns}
\begin{columns}\label{nomroot6ex}
    \cols /háhopìi/ & [ha˥oçiː˩˧] & \emh{hahópìi} & `sloth'
\end{columns}

\subsubsection{Overview}
An overview of the nominal stem structure of the Hapi language can be seen in table \ref{nstruco}.
In the first prefix slot, the possessive markers can be found, right before the root. Then, the augmentative and diminutive conjunct suffixes are placed right after the root. The case markers are located afterwards, only being interrupted by the number marker \emh{-(so)a}.
\begin{table}[p!]
    \centering
    \Rotatebox{90}{%
    \begin{tabular}{@{}ccccccclc@{}}
    \toprule
    \multicolumn{4}{c}{-1} & 0 & 1 & 2 & \multicolumn{1}{c}{3} & 4 \\ \midrule
    \multicolumn{4}{c}{Possession} & ROOT & Aug/Dim & Case & \multicolumn{1}{c}{Number} & Case \\ \midrule
     &  & \#V & \#C & \multirow{8}{*}{} & AUG & \multirow{8}{*}{see Section \ref{scase}} & \multicolumn{1}{c}{PL} & \multirow{8}{*}{see Section \ref{scase}} \\ \cmidrule(r){1-4} \cmidrule(lr){6-6} \cmidrule(lr){8-8}
    \multirow{2}{*}{1} & SG & \multirow{2}{*}{h́-} & há- &  & -áh- / -aóh- &  & \multicolumn{1}{c}{-(so)a} &  \\ \cmidrule(lr){2-2}
     & PL &  & hi- &  & DIM &  & \multicolumn{1}{c}{} &  \\ \cmidrule(lr){2-2} \cmidrule(lr){6-6}
    2 & SG & \multicolumn{2}{c}{ó-} &  & -xii- &  &  &  \\ \cmidrule(lr){2-2}
     & PL & oh- & o- &  & \multicolumn{1}{l}{\multirow{4}{*}{}} &  &  &  \\ \cmidrule(lr){2-2}
     & HON & pò- & pái- &  & \multicolumn{1}{l}{} &  &  &  \\ \cmidrule(lr){2-2}
    3 & SG & k̀- & kì- &  & \multicolumn{1}{l}{} &  &  &  \\ \cmidrule(lr){2-2}
     & PL & t̀- & tà- &  & \multicolumn{1}{l}{} &  &  &  \\ \bottomrule
    \end{tabular}}
    \caption{Nominal Stem Structure}
    \label{nstruco}
\end{table}

\subsection{Gender}\label{sgender}
While gender is not explicitly marked on nouns, it can be denoted by using a noun and the respective third person female or male pronoun, namely \emh{hoa} and \emh{kó}, juxtaposing it after the inflected noun root, as can be seen in examples (\ref{nomgender1}) and (\ref{nomgender2}).
\begin{example}
\label{nomgender1}
\bits hàòxa hoa
\gloss capybara TSF
\tr female capybara
\end{example}
\begin{example}
\label{nomgender2}
\bits hàòxa kó 
\gloss capybara TSM 
\tr male capybara
\end{example}
\subsection{Number}\label{snumber}
The Hapi language distinguishes two types of number marking on nouns, singular and plural, glossed as {\Sg} and {\Pl} respectively. While the singular is left unmarked, the plural is marked by the suffix \emh{-(so)a}, as can be seen in examples (\ref{ex:number1}) and (\ref{ex:number2}). Instead of the full form \emh{-soa}, the abbreviated form \emh{-s} may be used, as exemplified in (\ref{ex:number3}).\footnote{In this case the abbreviated plural marker \emh{-s} is morphophonologically debuccalized to /h/.}
\begin{examples}
\ex
\label{ex:number1}
\bits há- sáhpa -soa
\gloss FSPOSS arrow.head PL
\tr my arrowheads; my collection of arrowheads
\ex
\label{ex:number2}
\bits hóh -áh -a -óh
\gloss man AUG PL ERG
\tr (the) big men (had slain a snake...)
\ex 
\label{ex:number3}
\bits ó- koí -s -tah
\gloss SSPOSS hut PL PRL
\tr (we shall go) via your huts
\end{examples}
\subsection{Diminutives}\label{sdim}
To form a noun's diminutive form, the suffix \emh{-xii} is used. It cannot only used be for simple diminutives as in (\ref{ex:dim1}), but also is employed derivationally for certain noun>noun processes; for example, it may be used for the derivation of `grandson' from the word `son', or can be used pejoratively to derive nouns that are connotated with a certain derogatory sense, as in (\ref{ex:dim2}) and (\ref{ex:dim3}).
\begin{examples}
\ex
\label{ex:dim1}
\bits hápaah -xii 
\gloss dog DIM
\tr the little dog
\ex
\label{ex:dim2}
\bits pái- hósii -xii
\gloss SSHPOSS son DIM
\tr your\textsubscript{HON} grandson
\ex
\label{ex:dim3}
\bits páhsóoih -a -xii
\gloss bug PL DIM
\tr those pesky little bugs!
\end{examples}
\subsection{Augmentatives}\label{saug}
The augmentative form, which is marked by the suffix \emh{-áh} \Aug, encodes augmentative semantics, as well as derives kinship terms in the opposite direction as the diminutive, i.e. \emh{káixo} `mother' > \emh{káixoáh} `grandmother', as exemplified in (\ref{ex:aug1}) and (\ref{ex:aug2}). It may also encode some sort of honorific sense of praising, as can be seen in example (\ref{ex:aug3}).
\begin{examples}
\ex
\label{ex:aug1}
\bits táhaa -áh 
\gloss barrel AUG
\tr a big barrel
\ex
\label{ex:aug2}
\bits kápihoo -áh
\gloss father AUG
\tr (my) grandfather
\ex
\label{ex:aug3}
\bits hahópìi -áh
\gloss sloth AUG
\tr (Praise be!) The sloth god!
\end{examples}
There is also the augmentative suffix \emh{-aóh}, which may be employed in the same sense as \emh{-áh} \Aug. (\ref{ex:aug4}) gives an example for an augmentative using the second augmentative suffix.
\begin{examples}
\ex 
\label{ex:aug4} 
\bits ahitáh -aóh
\gloss iguana AUGA
\tr a huge iguana
\end{examples}
The two augmentative suffixes may also be stacked to increase the augmentative, praising meaning usually encoded by these morphemes. In this case, the suffix \emh{-áh} comes first, as can be seen in example (\ref{ex:aug5}).
\begin{examples}
\ex
\label{ex:aug5}
\bits sósikíí -áh -aóh
\gloss wooden.beam AUG AUGA
\tr an immense wooden beam
\end{examples}
\subsection{Possession}\label{spossession}
There are several types of possession in the Hapi language, which are presented in the following section. I will consider possessable nouns in section \ref{salien} and will discuss unpossessable nouns in section \ref{sinalien}. Concerning the different types of possession, an overview is given in this section. The Hapi language distinguishes between \textbf{(i)} possessable and \textbf{(ii)} unpossessable nouns. Unpossessable nouns can further be subdivided into \textbf{(iia)} indirect possessesable unpossessable nouns and \textbf{(iib)} inherently possessed unposessable nouns. 
\subsubsection{Possessable Nouns}\label{salien}
Most nouns in the language are possessable, i.e. can be marked for possession by the employment of personal markers which are prefixed to the root. I will discuss the employment of personal markers in \ref{spossmarker} and the interaction between possession and nominal classifier, discussing a construction called the possesive classifier construction in section \ref{snomclassifiersandpossession}
\paragraph{Personal Markers}\label{spossmarker}
There are personal marker which may be employed to encode possession on possessable nominals. Table \ref{tpossmarker} gives an overview of the personal marker paradigm.
\begin{table}[H]
\centering
\begin{tabular}{@{}cccc@{}}
\toprule
\multicolumn{2}{c}{\multirow{2}{*}{Person}} & \multicolumn{2}{c}{Morpheme} \\ \cmidrule(l){3-4} 
\multicolumn{2}{c}{} & \#V & \#C \\ \midrule
1 & SG & \multirow{2}{*}{h́-} & há- \\ \cmidrule(lr){2-2}
 & PL &  & hi- \\ \cmidrule(lr){2-2}
2 & SG & \multicolumn{2}{c}{ó-} \\ \cmidrule(lr){2-2}
 & PL & oh- & o- \\ \cmidrule(lr){2-2}
 & HON & pò- & pái- \\ \cmidrule(lr){2-2}
3 & SG & k̀- & kì- \\ \cmidrule(lr){2-2}
 & PL & t̀& tà- \\ \bottomrule
\end{tabular}
\caption{Possessive personal markers: Overview}
\label{tpossmarker}
\end{table}
Some examples for the usage of these personal markers can be found below. Example (\ref{ex:poss1}) and (\ref{ex:poss2}) showcase the employment of the '\#V floating tone' markers \emh{h́-} and \emh{t̀-}, while (\ref{ex:poss3}) exemplifies the usage of the honorific marker \emh{pái-}. As can be seen in example (\ref{ex:poss2}), the third person plural morpheme may also be used as an inter-clausal forth person pronoun. This is true for most employments of this pronoun, not only in its possessive form.
\begin{examples}
\ex
\label{ex:poss1}
\words háhoáta\MC2
\bits h́- ahoáta
\gloss FSPOSS bow
\tr my/our bow
\ex
\label{ex:poss2}
\words tàhoáta\MC2
\bits t̀- ahoáta
\gloss TPPOSS bow
\tr their bow/hisⱼ bow
\ex 
\label{ex:poss3}
\words páipáaxo\MC2
\bits pái- páaxo
\gloss SSHPOSS tendon
\tr your\textsubscript{HON} tendon (the one you extracted from the meat)
\end{examples}
\paragraph{Nominal Classifiers and Possession}\label{snomclassifiersandpossession}
There is also another special construction to express possession, using nominal classifiers and personal markers. This constructions is showcased in example (\ref{ex:poss4}). After the inflected noun, the according classifier is inserted, with the respective personal marker attached to it.\footnote{More on nominal classifiers in section \ref{sclassifier}} 
The classifier may also be moved to the front within a clause to focus the aspect of possession, as can be seen in example (\ref{ex:poss5}).
\begin{examples}
\ex
\label{ex:poss4}
\bits káhi kì- hóó
\gloss chicken.egg TSPOSS CL:egg  
\tr his chicken egg
\ex
\label{ex:poss5}
\words hóhóikaííxi\MC2 kói\MC2 hò\MC2 póóihi\MC2 xósóóxíkóa\MC4 hápaáh\MC2
\bits {hó$\sim$ hóika} =ííxi koí -∅ h- ò póói -i xósóó -xí -∅ =kóa hápaáh -h 
\gloss REDUP:build \textsc{\textup{s/a/o>o(se)}} hut ABS FSPOSS CL:anim:sg come DEP bite RPASTO ST DECL dog ERG 
\tr As he was once again building another hut, my dog came and bit (him)
\end{examples}
\subsubsection{Unpossessable Nouns}\label{sinalien}
There is a closed set of unpossessable nouns in the Hapi language, which either \textbf{(i)} can't be possessed directly, or \textbf{(ii)} can't be possessed at all, with a suppletive, possessable form being used for other means of possession. Those two types are called indirect possessable unpossessable nouns and inherently possessed unpossessable nouns. I will first consider indirect possessable unpossessable nouns in this section, just before presenting the different semantic aspects of inherently possessed unpossessable nouns, as well as showcase their suppletive possessable forms. 
The semantics of indirect possessable unpossessable nouns (henceforth indPUNs; cf. inherently possessed unpossessable nouns, inhPUNs) vary greatly within the language, and are summarized in table \ref{t:ccsgip}, where an overview of the most common cross-linguistic semantic groups in Hapi is given.
\begin{table}[H]
\centering
\captionsetup{justification=centering}
\begin{tabular}{@{}lll@{}}
\toprule
Semantic Group & Example & Possessability \\ \midrule
landscape features & rivers, mountains & indPUN \\
professions & hunters, shamans & indPUN \\
social relationships & neighbour, servant & indPUN \\
kinship terms & mother, brother & inhPUN \\
body parts & head, limbs & inhPUN \\
possessive attributes & age, name & inhPUN \\
mental states & fear, mind & (possessable) \\
part-whole relationships & side, top & (possessable) \\ \bottomrule
\end{tabular}
\caption{Common Cross-linguistic Semantic Groups and their Possessability in Hapi}
\label{t:ccsgip}
\end{table}
IndPUNs include geographical features, such as rivers, mountains or trees, as well as professions or persons, such as hunters and shamans. 
They all may be possessed via certain indirect possession strategies which are showcased in examples (\ref{ex:poss6}), (\ref{ex:poss7}) and (\ref{ex:poss8}). In these cases, the possession is semantically quite indirect, only encoding loose association, e.g. proximal distance to landscape features or the belonging to a family of professions and persons. 
\begin{examples}
\ex 
\label{ex:poss6}
\bits hòkaapa hi- táah
\gloss hunting.person FPPOSS CL:profession
\tr our hunter; the hunter that belongs to our family
\ex
\label{ex:poss7}
\bits xííkoh =à tàahoih
\gloss hill LINK FF
\tr my/our hill; the hill that is located near my/our hut
\ex
\label{ex:poss8}
\bits àkóóih xáa pòih
\gloss in.law COP SS.DAT
\tr your in-law
\end{examples}
In example (\ref{ex:poss6}), the construction explained in section \ref{snomclassifiersandpossession} is employed; this construction is not only used in these cases, but also, as explained in the referred section, to focus the possessed noun. In example (\ref{ex:poss7}), another construction is used. The linking disjunct affix \emh{=à} is suffixed to the head, while the respective emphatic pronoun form is inserted right after it.\footnote{Compare section \ref{spronouns} for a comprehensive overview of the Hapi pronoun system.}
In example (\ref{ex:poss8}), the copula is used to connect the head noun with the dative form of the possessing pronoun. This construction is rather sparsely used, while the other ones are employed quite often. 
\paragraph{Kinship Terms}
Kinship terms are one of the most often used inhPUNs in the Hapi language. As they're all inherently possessed by the first person singular, they each have a suppletive form that can be regularly possessed using the prefixes discussed in section \ref{spossmarker}. Table \ref{t:kinship} gives an overview of the most commonly used kinship inhPUNs and their suppletive possessable counterparts. 
\begin{table}[H]
\centering
\begin{tabular}{@{}lll@{}}
\toprule
inhPUN & Possessable Form & Meaning \\ \midrule
xòih & kàháa & (my) brother \\
kápihoo & káota & (my) father \\
haóxí & hósii & (my) son \\
káixo & xapisóo & (my) mother \\ \bottomrule
\end{tabular}
\caption{InhPUNs and Suppletive Forms (Kinship)}
\label{t:kinship}
\end{table}
Example (\ref{ex:poss9}), (\ref{ex:poss10}) and (\ref{ex:poss11}) showcase the usage of the suppletive forms. 
\begin{examples}
\ex
\label{ex:poss9}
\bits tài xàa haóxí -ò
\gloss DEM:PROX COP son CL:anim:sg
\tr This is my son
\ex
\label{ex:poss10}
\bits:* tài xàa ó- haóxí -ò
\gloss DEM:PROX COP SSPOSS son CL:anim:sg
\intended This is your son
\ex
\label{ex:poss11}
\bits tài xàa ó- hósii -ò
\gloss DEM:PROX COP SSPOSS son CL:anim:sg
\tr This is your son
\end{examples}
An extensive list of all inherently possessed nouns and their suppletive counterparts may be found in appendix A.
\subsection{Case}\label{scase}
In this section I will consider the language's case system. The Hapi case system is relatively complex, featuring a total of 10 cases, which can be subdivided into 3 core cases, 4 non-core cases and 3 relational cases. In section \ref{scorecase} I will discuss the language's core cases, in section \ref{sncorecase} I will present its non-core cases and in section \ref{srelcase} I will consider its relational cases. 
\subsubsection{Core Cases}\label{scorecase}
In the following section I will present the language's core cases. These cases are employed to mark core arguments such as S, A, P, D, T, and R for their appropriate role. In this context, the absolutive case marks the S of intransitive verbs, as well as the P of transitive and the T of ditransitive verbs. The ergative case marks the A of transitive verbs and the D of ditransitive verbs, while the dative case marks most commonly the R of a ditransitive verb, but also the second argument of a extended intransitive clause.\footnote{See section \ref{s:verbclass} for more information on extended intransitive clauses.} These cases are all marked in the first case slot.
\paragraph{Absolutive Case \emh{-∅} \textsc{abs}}
Employed to encode the S of intransitive clauses, the primary function of the absolutive case is showcased in example (\ref{ex:abs1}); the speaker informs the listener about an extraordinary warrior, who, earlier on that day, showed mercy for an enemy warrior. This is seen as a very unusual and strange action to be done by a warrior, hence the explicit statement of the fact. 
\begin{examples}
\ex \label{ex:abs1}
\words toíhoh\MC2 sáhaaxíkóa\MC4
\bits toíhoh -∅ sáhaa -xí -∅ =kóa 
\gloss warrior ABS show.mercy RPASTO ST  DECL 
\tr Earlier today, the warrior showed mercy
\end{examples}
Besides that, the absolutive case also marks the P of transitive clauses as in (\ref{ex:abs2}) and the T of ditransitive verb, thus making the Hapi language a indirective language. Example (\ref{ex:abs3}) gives an example for this tertiary function of the absolutive case.
In (\ref{ex:abs2}), the speaker talks about a young girl who ate an underripe mangaba fruit, and got stomach aches from it the day after. In (\ref{ex:abs3}), the speaker reports of a benevolent teacher, who gifts a student of his a present.
\begin{examples}
\ex \label{ex:abs2}
\words hoa sááhikóa\MC4 háiikáh\MC2
\bits hoa sáá -hi -∅ =kóa háiikáh -∅ 
\gloss TSF:AGENT eat RPASTT ST DECL type.of.fruit ABS
\tr She ate a mangaba fruit (yesterday)
\ex \label{ex:abs3}
\words kaósóaíh\MC2 kíih hóíakóa\MC3 xáháìih\MC2 xíooki\MC2
\bits kaósóa -íh kíih hóí -a =kóa xáháìi -h xíooki -∅
\gloss teacher ERG old give ST DECL student DAT gift ABS
\tr The old teacher gives a gift to a student
\end{examples}
\paragraph{Ergative Case \emh{-(V́)h} \textsc{erg}}
This case is used to mark the A of transitive clauses, as well as the D of ditransitive verbs. These two functions are exemplified in (\ref{ex:erg1}) and (\ref{ex:abs3}).
In (\ref{ex:erg1}), the neighbour of the speaker was taking the raw meat back from the fire place back to his hut, and this for unknown reasons.
\begin{examples}
\ex \label{ex:erg1}
\words àkóóihíh\MC2 xáh hóhiaka\MC3
\bits àkóóih -íh xáh hóhi -a =kóa 
\gloss  neighbour ERG raw.meat take ST DECL 
\tr (My) neighbour is taking the raw meat (with him)
\end{examples}
\paragraph{Dative Case \emh{-V̀Vh/-s} \textsc{dat}}
The dative case is used to mark the R of a ditransitive clause, as well as the extended argument of an extended intransitive clause. These two contexts are exemplified in (\ref{ex:abs3}) and (\ref{ex:dat1}). In (\ref{ex:dat1}), a child is playing with the mud.
\begin{examples}
\ex \label{ex:dat1}
\words kó sahóóhakóa\MC3 xóahasàxa\MC3
\bits kó sahóó -a =kóa xóaha -s -àxa
\gloss TSM:SUBJECT carve.in ST DECL mud DAT ONT
\tr He is scratching (something) in the mud
\end{examples}
\subsubsection{Non-Core Cases}\label{sncorecase}
The non-core cases are employed in the contexts of peripheral, non-core arguments, hence their name. There are four non-core cases, all of which are marked in the first case slot.
\paragraph{Locative Case \emh{-hóo} \textsc{loc}}
The locative case marks a noun for a locational meaning, i.e. the position in, from, or toward a certain object. Specifically, the locative case may bear a pure locational meaning, usually translated as `in, at', a lative meaning, which can be translated using the adposition `to(wards)', and an ablative meaning, i.e. `from, by'. These meanings are showcased in the examples (\ref{ex:loc1}), (\ref{ex:loc2}) and (\ref{ex:loc3}). In each of those, the speaker was asked to describe several contexts in which the verbs `be at', `go' and `come' were all used with the locative case. 
\begin{examples}
\ex \label{ex:loc1}
\words tàhokó hoihkóa\MC3 há pííháhóo\MC3
\bits tàhokó hoi -h =kóa há píí -áh -hóo
\gloss FAT:SUBJECT be.at FF DECL EGO river AUG LOC
\tr We are at the river
\ex \label{ex:loc2}
\words tàhokó háahkóa\MC3 há pííháhóo\MC3
\bits tàhokó háa -h =kóa há píí -áh -hóo
\gloss FAT:SUBJECT go FF DECL EGO river AUG LOC
\tr We go to the river
\ex \label{ex:loc3}
\words tàhokó póóihkóa\MC3 há pííháhóo\MC3
\bits tàhokó póói -h =kóa há píí -áh -hóo
\gloss FAT:SUBJECT come FF DECL EGO river AUG LOC
\tr We come from the river
\end{examples}
\paragraph{Perlative Case \emh{-tah} \textsc{prl}}\label{s:perlative}
The perlative case expresses that something is moved `through' or `along' the referent of the noun that is marked. It also marks the demoted A of passive constructions. The examples (\ref{ex:perl1}) and (\ref{ex:perl2}) demonstrate the case's usage. While (\ref{ex:perl1}) might not need any further explanation, (\ref{ex:perl2}) does. There, the speaker tells a narrative of a foolish young girl, who, at the end of the story, is eaten by a giant crab. This is a symbol for the punishment for her `unlawful' actions.
\begin{examples}
\newbaarucmd{\cl}{\baarujuncture{\texttt{==}}}
\ex \label{ex:perl1}
\words tàahopí há hááháah\MC3 ókoíhtah\MC4
\bits tàahopí há háá- háa -h ó- koí -s -tah 
\gloss FAS:SUBJECT EGO JUSS go FF SSPOSS hut PL PRL
\tr We shall go via your huts
\ex \label{ex:perl2}
\words hoó asááhàòkóapò\MC5 xáohaóhtah\MC3
\bits hoó a- sáá -hàò =kóa \cl pò xáoh -aóh -tah
\gloss TSF:PATIENT PASS eat DPASTO DECL INFER crab AUGA PRL
\tr She was eaten by a huge crab (they told me)
\end{examples}
\paragraph{Privative Case \emh{-sáahi} \textsc{pvt}}
The privative case expresses the acting of a verb `without' a certain noun, more or less yielding a construction of a `\textsc{noun}-less' meaning. Together with the copula marker \textit{xàa}, the semantics of `to not have \textsc{noun}' may be achieved. This and the more general meaning of the privative case are shown in examples (\ref{ex:pvt2}) and (\ref{ex:pvt1}). (\ref{ex:pvt1}) tells us about a young, inexperienced hunter who doesn't quite know how to separate meat from fat elegantly. In (\ref{ex:pvt2}), the speaker utters this sentence in the context of a hunt, in which another hunter has slain a green iguana, whereas the speaker has not.
\begin{examples}
\ex \label{ex:pvt1}
\words kó sahóóhikóa\MC3 tíahsáahi\MC2
\bits kó sahóó -hi =kóa \\ tíah -sáahi
\gloss TSM:SUBJECT separate.meat.from.fat RPASTT DECL care PVT
\tr He separated meat from fat carelessly
\ex \label{ex:pvt2}
\words ahitáhsáahi\MC2 xàa tà
\bits ahitáh -sáahi xàa tà
\gloss green.iguana PVT COP FS
\tr (Unlike you,) I do not have a green iguana
\end{examples}
\paragraph{Semblative Case \emh{-haixá} \textsc{smb}}
The semblative case denotes the similarity of the subject to the marked noun. This function is explained in example (\ref{ex:smb1}). There, the speaker narrates an event from his childhood, in which a sick stranger arrived at the village; his condition is described in (\ref{ex:smb1}).
\begin{examples}
\ex \label{ex:smb1}
\words haxápií hóahtóhikóa\MC3 tísoohaixá\MC2
\bits haxápií hóah -tóhi =kóa tísooh -haixá
\gloss stranger vomit DPASTO DECL frog SMB 
\tr The stranger vomited like a frog
\end{examples}
\subsubsection{Relational Cases}\label{srelcase}
In this section I will consider the relational cases of the Hapi language. These cases are employed to specify a certain relational sense, i.e. they mark a noun for their spatial relationship with another noun. As will be discussed in section \ref{scasestacking}, these cases may be stacked with other non-core cases to yield more specified semantics. These cases are simply named after their English semantic equivalents;
\paragraph{`on top of' \emh{-(V̀)xa} \textsc{on.top} }
The `on top of' case primarily\footnote{Another usage is sketched out in section \ref{s:trans}} marks a noun's position to be roughly `on top of' another noun, as can be seen in example (\ref{ex:ontop1}). (\ref{ex:ontop2}) shows, how a construction of the shape `x-\textsc{on.top} y' can indicate possession in the form of `y has x'.
\begin{examples}
\ex \label{ex:ontop1}
\words haípi haahakóa\MC3 píihàxa\MC2
\bits haípi haah -a =kóa píihà -xa
\gloss soup cook ST DECL fire.place ONT
\tr The soup is cooking above the fire place
\ex \label{ex:ontop2}
\words kihaòhxa\MC2 póxihoo
\bits kihaòh -xa póxihoo
\gloss type.of.fruit ONT SS
\tr You/youₚₗ have a tucumã fruit
\end{examples}
\paragraph{`below' \emh{-(t)aó} \textsc{below}}
The `below' case marks a noun's position to be roughly `below' another noun, as can be seen in example (\ref{ex:below1}).
\begin{examples}
\ex \label{ex:below1}
\words tàhòhì tísooh hoí xasíítaó\MC2
\bits tàhòhì tísooh hoí xasíí -taó
\gloss poison frog COP leaf BEL
\tr The poison dart frog sits under the leaf
\end{examples}
\paragraph{`next to' \emh{-(h)aí} \textsc{next.to}}
The `next to' case marks a noun's position to be roughly `next to' another noun, as can be seen in example (\ref{ex:nextto1}). There, the speaker reports of a girl that was, defying the traditional Hapi culture, eating a soup next to her father, even though said traditions don't allow members of either gender to eat in the proximity of each other.
\begin{examples}
\ex \label{ex:nextto1}
\words pihih tááhtóhikóa\MC3 haípi kìkáotahaí\MC3
\bits pihih tááh -tóhi =kóa haípi kì- káota -haí
\gloss girl drink DPASTO DECL soup TSPOSS father NEXT
\tr The girl was drinking soup next to her father
\end{examples}
\subsubsection{Case
Stacking}\label{scasestacking}
In the Hapi language, second slot cases may be stacked with first slot cases to yield specific semantics, which are shown in table \ref{t:casestacking}. Note that the privative and semblative cases do not allow any relational combination.
\begin{table}[H]
\begin{tabular}{@{}ccccc@{}}
\toprule
\multirow{2}{*}{} & \multicolumn{3}{c}{Locative}                        & \multirow{2}{*}{Perlative} \\ \cmidrule(lr){2-4}
                  & Locational        & Lative         & Ablative       &                            \\ \midrule
'on top of'       & focal 'on top of' & 'to above'     & 'from above'   & 'via above'                \\
'below'           & foc. 'below'      & 'to below'     & 'from below'   & 'via below'                \\
'next to'         & foc. 'next to'    & 'to (next to)' & 'from next to' & 'via next to'              \\ \bottomrule
\end{tabular}
\caption{Case Stacking Combinations}
\label{t:casestacking}
\end{table}
The relational-locational combinations that are shown in the table are 'focal' variants of the normal, uncombined relational cases and contrast with them in a such a way that the focal versions are more particular. An example for this is given in (\ref{ex:casestacking1}), where, after a child is told to put a smaller stone onto a bigger one, the child asks whether it has done so correctly. The reply is (\ref{ex:casestacking1}).
\begin{examples}
\ex \label{ex:casestacking1}
\words kií, haò kaíi hàohóoxa\MC3
\bits kií haò kaíi haò -hóo -xa
\gloss no rock COP rock LOC ONT
\tr No, the rock must be (exactly) on top of the (other) rock
\end{examples}
%Maybe: examples for every cell in the table? Perhappu a bit too much :pensive: 
\subsection{Conclusion}
In the preceding section, I have discussed the structure of the nominal root in section \ref{snomroot}, the ways of encoding gender, number, diminutives and augmentatives in the sections \ref{sgender}, \ref{snumber}, \ref{sdim} and \ref{saug} respectively, have considered possession in its comprehensive aspects in section \ref{spossession} and have presented the Hapi case system in section \ref{scase}.
\section{Noun phrase structure}\label{snounphrase}
The structure of the Hapi noun phrase, while not as complex as the verb phrase's, is still worthy of a comprehensive discussion. In the following section I will first consider the order of noun phrase elements in section \ref{snounorder}, then the language's pronoun system in section \ref{spronouns}, before moving on to the extensive classifier system in section \ref{sclassifier}. Afterwards I will lay out different parts of the remaining noun phrase, namely, demonstratives in section \ref{sdemonstrative}, quantifiers and numerals in section \ref{squantifier}, concluding with attributes such as adjectives and possessives in section \ref{sattributes}. 
\subsection{Order of noun phrase elements}\label{snounorder}
The elements of the noun phrase appear in a specific order, as schematized in example (\ref{ex:nounphraseorder}).
\begin{examples}
\ex \label{ex:nounphraseorder}
(QUANT) - (DEM) - (N) - head -  (PSR) - (ADJ) 
\end{examples}
Thus, quantifiers with the respective classifiers (\ref{ex:orderquant}), demonstratives (\ref{ex:orderdem}), and nouns modifying the head (\ref{ex:ordernoun}) all precede the noun, while possessors (\ref{ex:orderpsr}) and adjectives (\ref{ex:orderadj}) follow it.
\begin{examples}
\ex \label{ex:orderquant}
\bits xasó -soo kìah -a
\gloss six CL:fingers finger PL
\tr six fingers
\ex \label{ex:orderdem}
\bits hao kahíí
\gloss DEM:DIST pig
\tr that pig
\ex \label{ex:ordernoun}
\bits xòxíi haíka
\gloss leaves type.of.palm
\tr maripa palm leaves
\ex \label{ex:orderpsr}
\bits sosáai =à kápihoo
\gloss cat LINK father
\tr my father's cat
\ex \label{ex:orderadj}
\bits hati tóhaki
\gloss servant foolish
\tr the foolish servant
\end{examples}
\subsection{Pronouns}\label{spronouns}
The Hapi pronominal system distinguishes on a first level between so-called independent and possessive pronouns. The independent pronouns take the role of S, A and O, as well as an instrumental usage. They ca be subdivided into ergative, neutral and accusative pronouns depending on their morphosyntactic alignment. The system also makes the distinction between person and number, as well as clusivity in first person pronoun. There is also a dedicated honorific pronoun \emh{hií} and a third person plural pronoun which can be used as fourth person pronoun.
The set of possessive pronouns are used in an independent or strong manner, i.e. not with a determinative sense, but rather as a true pronominal, cf. 'my' (determinative) and 'mine' (independent).
\subsubsection{Independent pronouns}
An overview of independent non-possessive pronouns is schematized in table (\ref{t:indeppron}).
\begin{table}[H]
\begin{centering}
    \begin{tabular}{@{}llllll@{}}
    \toprule
    \multicolumn{1}{c}{} & A                    & S                    & P                    & Dat              & Instr                  \\ \midrule
    Ergative:            & \multicolumn{1}{c}{} & \multicolumn{1}{c}{} & \multicolumn{1}{c}{} &                  & \multicolumn{1}{c}{}   \\
    1S                   & \textit{tàah}        & \textit{tà}          & \textit{tà}          & \textit{tàh}     & \textit{tàsóo}         \\
    1+2                  & \textit{tàahopí}     & \textit{tàhopí}      & \textit{tàhopí}      & \textit{tàhopìh} & \textit{-}             \\
    2S                   & \textit{páó}         & \textit{pói}         & \textit{pói}         & \textit{pòih}    & \textit{pósòo}         \\
    3P or 4              & \textit{koìh}        & \textit{xah}         & \textit{xah}         & \textit{xàh}     & \textit{xasóó}         \\
    Neutral:             &                      &                      &                      &                  &                        \\
    1+3                  & \textit{tàhokó}      & \textit{tàhokó}      & \textit{tàhokó}      & \textit{tàhokòh} & -                      \\
    2P                   & \textit{póh}         & \textit{póh}         & \textit{póh}         & \textit{pòh}     & -                      \\
    Accusative:          &                      &                      &                      &                  &                        \\
                         & \textit{kó}          & \textit{kó}          & \textit{kòà}         & \textit{kòh}     &                        \\
                      3S & \textit{hoa}         & \textit{hoa}         & \textit{hoó}         & \textit{hòah}    & \textit{hisóo}                  \\
                         & \textit{íí}          & \textit{íí}          & \textit{kòà}         & \textit{ìih}     &                        \\
    2Hon                 & \textit{hií}         & \textit{hií}         & \textit{pói}         & \textit{hìih}    & \textit{pásoò}         \\ \bottomrule
    \end{tabular}
    \caption{Independent non-possessive pronominal system
    }
    \label{t:indeppron}
\end{centering}
\end{table}
\paragraph{Ergative pronouns}
The ergative pronouns, as can be inferred from their name, have merged the forms for S and P, while the A form remained distinct. They include the first and second person singular, first person plural inclusive and third person plural pronouns. Some usages of ergative pronouns are found in examples (\ref{ex:ergpron1}) - (\ref{ex:ergpron3}).
\begin{examples}
\newbaarucmd{\cl}{\baarujuncture{\texttt{==}}}
\ex \label{ex:ergpron1}
\words tà hóíkatah\MC2 háahoò\MC4 xóatiho\MC3 haíhosoahò\MC3 há
\bits tà hóíka -tah háa -h -o \cl ò xóati -h -o haího -soa \cl ò há 
\gloss FS:SUBJECT big.tree PRL go FF FUT SEQ search.for FF FUT berry PL SEQ EGO
\tr I'll be going via the big tree, and then (I) will search for berries (there)
\ex \label{ex:ergpron2}
This section is work-in-progress
\ex \label{ex:ergpron3}
This section is work-in-progress
\end{examples}
\paragraph{Neutral pronouns}
The neutral or direct pronouns have a single form for all core functions. They consist of the first person plural exclusive and the second person plural pronouns. Some examples are given in (\ref{ex:neutrpron1}) and (\ref{ex:neutrpron2}).
\begin{examples}
\ex \label{ex:neutrpron1}
This section is work-in-progress
\ex \label{ex:neutrpron2}
This section is work-in-progress
\end{examples}
\paragraph{Accusative Pronouns}
The set of accusative pronouns that are employed in the Hapi language possess a form for A and S, and another, distinct form for P. They feature the third person singular pronouns, as well as the second person honorific pronoun. Some example sentences using these pronouns are given in (\ref{ex:accpron1}), (\ref{ex:accpron2}) and (\ref{ex:accpron3}).
\begin{examples}
\ex \label{ex:accpron1}
This section is work-in-progress
\ex \label{ex:accpron2}
This section is work-in-progress
\ex \label{ex:accpron3}
This section is work-in-progress
\end{examples}
\paragraph{Dative Pronouns}
The set of dative pronouns may be used in the same way the dative case is employed. It marks the recipient of a ditransitive clause, as well as the extended argument of an extended intransitive clause. It is also used in some possessive constructions. The examples (\ref{ex:datpron1}) and (\ref{ex:datpron2}) showcase the usage of these pronouns.
\begin{examples}
    \ex \label{ex:datpron1}
    This section is work-in-progress
    \ex \label{ex:datpron2}
    This section is work-in-progress
    \end{examples}
\paragraph{Instrumental pronouns}
The majority of the non-possessive pronouns have a suppletive form that can be used to denote the semantic instrument of a verb. Only the first and second person plural pronouns do not have an additional instrumental form. Some examples for the instrumental usage of these forms are given in (\ref{ex:instrpron1}) and (\ref{ex:instrpron2}).
\begin{examples}
\ex This section is work-in-progress. \label{ex:instrpron1}
\ex This section is work-in-progress. \label{ex:instrpron2}
\end{examples}
When using the first person singular, second person singular and second person honorific instrumental pronoun, another reading may be achieved, namely, an intensification of the salience of the subject or agent. The examples in (\ref{ex:instrpron3}) and (\ref{ex:instrpron4}) showcase the usage thereof.
\begin{examples}
\ex \label{ex:instrpron3}
\words áahíkóa\MC4 `tàah kohaíkoahi\MC2 kiihikóa\MC4 há tàsóo' ko
\bits áa -hi -∅ =kóa tàah kohaíkoa -i kii -hi -∅ =kóa há tàsóo ko
\gloss say RPASTT ST DECL FS:AGENT dig.a.hole  DEP AUX:PERF RPASTT ST DECL  EGO FS:INSTR QUOTATIVE
\tr He said: ``I dug a hole all by myself''. 
\ex This section is work-in-progress. \label{ex:instrpron4}
\end{examples}
\subsubsection{Possessive pronouns}
The independent possessive pronouns are given in table (\ref{t:posspron}).
\begin{table}[H]
\begin{centering}
\begin{tabular}{@{}lccc@{}}
\toprule
\multicolumn{1}{c}{} & SG         & PL        & HON   \\ \midrule
1                    & háhi       & hí        & -     \\
2                    & \multicolumn{2}{c}{óó} & páihi \\
3                    & kìhi       & tài       & -     \\ \bottomrule
\end{tabular}
\caption{Independent possessive pronouns}
\label{t:posspron}
\end{centering}
\end{table}
Some examples for the usage of these are presented in (\ref{ex:posspron1}) and (\ref{ex:posspron2}).
\begin{examples}
\ex \label{ex:posspron1}
\words tài hati íxàa\MC2 há óóhò?\MC2
\bits tài hati í- xàa há óó -ò
\gloss DEM:PROX servant INTRG COP EGO SSPOSS CL:anim:sg
\tr Is that servant yours?
\ex \label{ex:posspron2}
\words ò kahaóó xàa kìhiò[...]\MC2
\bits ò kahaóó xàa kìhi -ò
\gloss and uncle COP TSPOSS CL:anim:sg
\tr And it was his uncle [...]
\end{examples}
\subsubsection{Demonstrative Pronouns}
The set of demonstrative pronouns consists of N forms, which are shown in table \ref{t:dempron}. Note the similarity between these and the demonstrative adjectives presented in section \ref{sdemonstrative}.

\begin{table}[H]
\centering
\begin{tabular}{@{}llll@{}}
\toprule
     & VIS        & NVIS                     & INANIM                                      \\ \midrule
PROX & tài 'this' & -                        & \multirow{2}{*}{ho 'this/that' (inanimate)} \\
DIST & hao 'that' & hih 'that' (not visible) &                                           \\ \bottomrule
\end{tabular}
\caption{Demonstrative Pronominal System}
\label{t:dempron}
\end{table}

Some cases for the usage of these pronouns are exemplified in (\ref{ex:dempron1}), (\ref{ex:dempron2}) and (\ref{ex:dempron3}).
\begin{examples}
\ex \label{ex:dempron1}
\words tài xàa áaxíhao\MC4 há kóhii kó xàa kaai hóhò\MC2
\bits tài xàa áa -xí -∅ -hao há kóhii kó xàa kaai hóh -ò  
\gloss DEM:PROX COP say RPASTO ST DEM EGO C TSM:SUBJECT COP sick man CL:anim:sg
\tr This (pointing at man) is the man who said [of himself] that he was sick
\ex \label{ex:dempron2}
\words hákapióó áahaokóa\MC4 ho
\bits hákapióó áa -a -o =kóa ho
\gloss nobody say ST FUT DECL that
\tr Nobody is going to say that
\ex \label{ex:dempron3}
\words hih xàa óxííhò\MC2
\bits hih xàa óxíí -ò 
\gloss  DEM:NVIS COP child CL:anim:sg
\tr That (pointing at approximate direction) was a child [not a rock]
\end{examples}
\subsubsection{Emphatic pronouns}
Besides the set of independent, possessive and demonstrative pronouns, there is also a system of emphatic pronouns which are used in certain contexts. The set of those pronouns is depicted in table \ref{t:emphpron}. Note how there is neither a clusivity nor a number nor a honorific distinction within the pronouns of this set.
\begin{table}[H]
    \begin{centering}
    \begin{tabular}{@{}lc@{}}
    \toprule
    \multicolumn{1}{c}{} & SG/PL    \\ \midrule
    1                    & tàahoih  \\
    2                    & póxihoo  \\
    3                    & hóaxáko \\ \bottomrule
    \end{tabular}
    \caption{Emphatic Pronouns}
    \label{t:emphpron}
    \end{centering}
\end{table}
These pronouns are used in N major contexts: \textbf{a)} in possessive constructions using the disjunct affix \emh{=à} (\ref{ex:emphpron1}), \textbf{b)} as emphatic pronouns (\ref{ex:emphpron2}). 
\begin{examples}
    \ex \label{ex:emphpron1}
    \words xoxííhà\MC2 tàahoih
    \bits xoxíí =à tàahoih
    \gloss palm.fiber LINK FF
    \tr my/our palm fiber
    \ex \label{ex:emphpron2}
    \words hóaxáko taíí kohaa toihaka\MC3 
    \bits hóaxáko taíí kohaa toih -a =ka 
    \gloss TT glass hole open ST DECL
    \tr It is him/her who is opening the window
\end{examples}
\subsection{Classifiers}\label{sclassifier}
In the following section I will discuss the language's extensive classifier system. I will start by considering an overview as well as a brief explanation of the different classifiers; afterwards I will clarify the syntactic distribution of these classifiers. 
\subsubsection{Overview and Features}
The different features all can be categorized in the values they can take. The first distinction that is to be made outside of the feature matrix is the MATERIAL distinction. This distinction that is made distinguishes animate, inanimate and abstract nouns, as can be inferred from table \ref{t:clmaterial}.
\begin{table}[H]
\begin{centering}
\begin{tabular}{@{}llll@{}}
\toprule
   & Animate & Abstract            & Inanimate                           \\ \midrule
SG & ò      & \multirow{2}{*}{ki} & \multirow{2}{*}{\textit{(see tables below)}} \\
PL & haí     &                     &                                     \\ \bottomrule
\end{tabular}
\caption{MATERIAL distinction}
\label{t:clmaterial}
\end{centering}
\end{table}

The other features include SHAPE (long, flat and round\footnote{Also referred to as saliently one/two/three-dimensional.}), SIZE (big, small), QUANTA (singular, plural, bunch of, basketful), RIGIDITY (flexible, rigid, brittle, non-discrete) and LOCATION (extended vertically, ext. horizontally, parallel objects, objects in a row). Table \ref{tab:longclass} gives an overview of some saliently one-dimensional classifiers in Hapi. A full list of classifiers can be found in appendix B.

\begin{table}[H]
    \begin{tabular}{lll>{\bfseries}l>{\itshape}p{5cm}}
        \toprule
        \multicolumn{5}{l}{Saliently 1D - Long} \\
        \midrule
        Big   & Flexible     &          & hó    & ropes, ladders, (long) vines          \\
              & Rigid        & Singular & ká    & trunks/pole-like objects that have fallen over    \\
              &              &          & saxá  & palm tree trunk, pole-like objects                 \\
              &              &          & hió   & pillars (of a house), supporting beams             \\
        \midrule
        Small & Flexible     &          & xái   & short hair, short ropes                            \\
              & Rigid        & Singular & aáxi  & single banana                                      \\
              &              &          & óah   & tubes, hollow objects, bones, flutes              \\
              &              & Plural   & hosó  & tubes, hollow objects, bones (plural)              \\
              & Brittle      & Singular & kaata & small branches of trees which are easily breakable \\
              & Non-Discrete &          & xoxi  & liquids in a long container, water (in a bottle)   \\
              &              &          & xóhii & viscous liquids \\
        \bottomrule
        \end{tabular}
\caption{Some Saliently 1D Classifiers}
\label{tab:longclass}
\end{table}
\subsubsection{Syntactic Distribution}
In the following sub-section I will discuss the syntactic distribution of the classifiers presented above. I will not mention the quantifier-head concordance that is present in the Hapi language, as it already has been considered in this same section.
\paragraph{Classifier possession}
A prominent usage of classifiers in the Hapi language is classifier possession. This construction was already shown in example (\ref{ex:poss4}), but shall be reiterated here. The examples (\ref{ex:clsyndistr1}) - (\ref{ex:clsyndistr3}) show the general structure of such constructions. Right after the possessee, the classifier follows its head, before the possessor is introduced to the phrase. 
\begin{examples}
\newbaarucmd{\cl}{\baarujuncture{\texttt{==}}}
\ex \label{ex:clsyndistr1}
\words haí hápaáh\MC2 ó táosiihó haxápií\MC2 xósóóhió\MC4 xah sóxiihió[...]\MC4 
\bits haí hápaáh -h ó táosiihó haxápií -∅ xósóó -hi -∅ \cl ó xah sóxii -hi -∅ \cl ó 
\gloss thus dog ERG CL:anim male.cousin stranger ABS bite RPASTT ST SEQ FOURTH scream RPASTT ST SEQ
\tr (And) thus the dog of my (male) cousinᵢ bit the strangerⱼ, and heᵢ screamed, [and ...]
\ex \label{ex:clsyndistr2}
\words pasío hosi kàaha
\bits pasío hosi kàaha 
\gloss blow.gun CL:thin.pole.like man's.name
\tr Kàaha's blow gun
\ex \label{ex:clsyndistr3}
\words papáxi xào póhsa
\bits papáxi xào póhsa
\gloss mouth CL:facial.body.parts jug
\tr jug's mouth
\end{examples}
\paragraph{Predicate Nominals}
Predicate nominals also agree with their copular subject by means of being suffixed the adequate classifier, as can be seen in example (\ref{ex:clsyndistr4}), (\ref{ex:clsyndistr5}) and (\ref{ex:clsyndistr6}). 
\begin{examples}
\ex \label{ex:clsyndistr4}
\words tài kohi xàa toohóósaxá\MC2
\bits tài kohi xàa toohóó -saxá
\gloss DEM:PROX trunk COP type.of.palm CL:trunks
\tr This trunk is of a peach palm
\ex \label{ex:clsyndistr5}
\words xóah xàa hòkaapaò\MC2
\bits xóah xàa hòkaapa -ò
\gloss man's.name COP hunter CL:anim:sg
\tr Juan is a hunter
\ex \label{ex:clsyndistr6}
\words hao táhiáki xàa páihikahía\MC2
\bits hao táhiáki xàa páihi -kahía
\gloss DEM hammock COP SSHPOSS CL:hanging
\tr This hammock is yours\textsubscript{HON}
\end{examples}
\subsection{Demonstrative Adjectives}\label{sdemonstrative}
The demonstrative adjectives of the Hapi language are very similar, if not identical to those demonstratives described in \ref{spronouns}. A short overview is given in table \ref{t:demdet}.
\begin{table}[H]
\centering
\begin{tabular}{@{}llll@{}}
\toprule
     & VIS        & NVIS                     & INANIM                                       \\ \midrule
PROX & tài 'this' & -                        & \multirow{2}{*}{hao 'this/that' (inanimate)} \\
DIST & hao 'that' & híi 'that' (not visible) &                                              \\ \bottomrule
\end{tabular}
\caption{Demonstrative Adjectives Overview}
\label{t:demdet}
\end{table}
Some examples for the usage of these are given in (\ref{ex:demdet1}) and (\ref{ex:demdet2}).
\begin{examples}
\ex \label{ex:demdet1}
\words hao hahópìi xàa tàiha
\bits hao hahópìi xàa tàiha
\gloss DEM:DIST sloth COP there
\tr There is that sloth [which we saw earlier]
\ex \label{ex:demdet2}
\words híi tukáh háhihahíaka\MC3
\bits híi tukáh háhihahí -a =ka 
\gloss DEM:NVIS toucan sing ST DECL
\tr That toucan is singing
\end{examples}
\subsection{Quantifiers and Numerals}\label{squantifier}
\subsubsection{Quantifiers}
Non-numeral quantifiers precede the noun together with the adequate classifier. Examples can be seen in (\ref{ex:quant1}), (\ref{ex:quant2}) and (\ref{ex:quant3}).
\begin{examples}
\ex \label{ex:quant1}
This section is work-in-progress
\ex \label{ex:quant2}
This section is work-in-progress
\ex \label{ex:quant3}
This section is work-in-progress
\end{examples}
\subsubsection{Numerals}
Numeral quantifiers also precede the modified noun, together with the adequate classifier. More information on numerals is given in chapter \ref{c:numerals}.  The position of numerals within the noun phrase is exemplified in (\ref{ex:numeral1}), (\ref{ex:numeral2}), and (\ref{ex:numeral3}).
\begin{examples}
\ex \label{ex:numeral1} 
\words hao akahííopa ahóikahàòkóa\MC5 hikoíípahahì\MC2 \\ {apíhaiha koha}\MC2 hakó
\bits hao akahííopa a- hóika -hàò -∅ =kóa  hikoíípaha -hì {apíhaiha koh} -a hakó
\gloss DEM bridge PASS build DPASTO ST DECL ten CL:time year PL ago
\tr That bridge was built ten years ago
%\source \parencite[36]{smoyd1308}
\ex \label{ex:numeral2}
\words tóhò\MC2 óxíísoa\MC2
\bits tóh -ò óxíí -soa
\gloss three CL:anim:sg child PL
\tr three children
\ex \label{ex:numeral3}
\words posá\MC2 aíxia\MC2 xàa tàh
\bits po -sá aíxi -a xàa tàh
\gloss nine CL:banana.trunks banana.plant PL COP FS:DAT
\tr I have nine banana tree plants
\end{examples}
\subsection{Attributes}\label{sattributes}
\subsubsection{Adjectives}
Adjectival attributes follow the noun. More information on adjectives is given in chapter \ref{c:adjectivesadverbs}. The order of adjectival elements in the noun phrase is given below in the examples (\ref{ex:adj1}) and (\ref{ex:adj2}).
\begin{examples}
\ex \label{ex:adj1}
This section is work-in-progress
\ex \label{ex:adj2}
This section is work-in-progress
\end{examples}
\subsubsection{Possessives}
The main strategy for possessives in the Hapi language is depicted in the examples (\ref{ex:nposs1}) and (\ref{ex:nposs2}). There, the linking disjunct affix \emh{=à} is attached to the head, followed by the possessor in the absolutive case. 
\begin{examples}
\baarucols2
\ex \label{ex:nposs1}
\words óíhtià\MC2 sóasih
\bits óíhti =à sóasih
\gloss monkey LINK person
\tr the person's monkey
\ex \label{ex:nposs2}
\words kókókóà\MC2 haxápií
\bits kókókó =à haxápií
\gloss gun LINK stranger
\tr the stranger's gun
\end{examples}
Another strategy to mark possession on nouns is the so-called copular strategy. There, the possessor is put in the dative case and is connected to the possessee by the copula \emh{xàa}. This process can be seen in the examples (\ref{ex:nposs3}) and (\ref{ex:nposs4}).
\begin{examples}
\ex \label{ex:nposs3}
This section is work-in-progress
\ex \label{ex:nposs4}
This section is work-in-progress
\end{examples}
\subsubsection{Noun modifying}
The final type of attribute in the Hapi language is noun juxtaposition. A noun may modify another noun in which case the modifying noun precedes the head noun. (\ref{ex:nounmod1}) and (\ref{ex:nounmod2}) show examples thereof.
\begin{examples}
\baarucols2
\ex \label{ex:nounmod1}
\bits tahai xohoá
\gloss goat milk
\tr goat milk
\ex \label{ex:nounmod2}
\bits hiiho kohkí
\gloss sheep fur
\tr wool
\end{examples}
\subsection{Conclusions}
In the preceding chapter I have considered the structure of the nominal word and the nominal phrase. I have presented the internal structure of the nominal root, demonstrated the morphological categories that are marked on the noun, namely, number, diminutive and augmentative, possession and case, and have discussed the formation of gender-specified nouns by juxtaposition. After a brief consideration of the order of noun phrase elements, I then moved on to the pronominal system, schematized the extensive set of classifiers of the language, and presented further any other noun modifiers such as demonstratives, quantifiers and numerals, adjectives, possessives and noun modifying nominals, all in the context of the noun phrase.
\chapter{The Verb and the Verb Phrase}
This chapter discusses the verb and the verb phrase in the Hapi language. In the first part of this chapter, I will discuss verbal morphology, considering the structure of the verbal stem and its conjuncts, disjuncts and clitics. In section \ref{sverbalroot} I will consider the structure of the verbal root; then I will move through the verb slot by slot, starting with the mode affix in section \ref{svmode}. Subsequently I will discuss valency in section \ref{svvalency}, past tenses in \ref{svpast}, person agreement in section \ref{svperson} and the future tense in section \ref{svfut}. In the second part of this chapter, I will explain the verb phrase structure, beginning with the order of verb phrase elements in section \ref{svorder}. Then, I will consider the relationship of second-position particles within the verb phrase in section \ref{svparticles}. Finally, I will present the role of adverbs in section \ref{svadverbs}.
\section{Verbal Morphology}
In this section, the structure of Hapi verbs is examined, focussing on the grammatical categories encoded by morphology, i.e. mode, valency, tense and person.
\subsection{Verbal Root}\label{sverbalroot}
The structural properties of the verbal root include a set of prefixes and suffixes for the expression of mode, valency, tense and person. 
\subsubsection{Internal stucture of verbal roots}
In the follwowing section I will consider the internal phonological structure of verbal roots, analyzing the syllabic composition of said roots. 
%% Monosyllabic: 13
%% Disyllabic: 25
%% Trisyllabic and polysyllabic roots: 4
There is a set of \textit{13} monosyllabic, \textit{26} disyllabic and \textit{4} trisyllabic or polysyllabic roots in the corpus of \textit{42} Hapi verbs.
\subsubsection{Overview}
An overview of the verbal stem structure of the Hapi language can be seen in table \ref{t:vstruc}. In the first slot, the mode conjunct prefixes are found; in the second slot, the valency conjunct prefixes are located. After the root, the past, person, future and causative markers can be found. Finally, the declarative disjunct marker is located in the final slot.
\begin{table}[p!]
    \centering
    \Rotatebox{90}{%
    \begin{tabular}{@{}llllllllll@{}}
    \toprule
    \multicolumn{2}{l}{-2}   & -1    & 0    & \multicolumn{2}{l}{1}    & 2           & 3      & 4     & 5    \\ \midrule
    \multicolumn{2}{l}{Mode} & Val   & ROOT & \multicolumn{2}{l}{Past} & Person      & Future & Caus  & Decl \\ \midrule
    INDIC       & ∅-         & ANTIP &      & RPAST1       & -xí       & see Section \ref{svperson} & -o     & CAUS1 & =kóa \\ \cmidrule(r){1-1} \cmidrule(lr){3-3} \cmidrule(lr){5-5} \cmidrule(lr){9-9}
    INTRG       & í-         & h(V)- &      & RPAST2       & -hi       &             &        & -áh   &      \\ \cmidrule(r){1-1} \cmidrule(lr){5-5}
    JUSS        & háá-       & PASS  &      & DPAST1       & -tóhi     &             &        & CAUS2 &      \\ \cmidrule(r){1-1} \cmidrule(lr){3-3} \cmidrule(lr){5-5} \cmidrule(lr){9-9}
                &            & a-    &      & DPAST2       & -hàò      &             &        & -kó   &      \\ \bottomrule
    \end{tabular}}
    \caption{Verbal Stem Structure}
    \label{t:vstruc}
\end{table}
\subsubsection{Verb Classes}\label{s:verbclass}
There are six classes of verbs in the Hapi language. Firstly, there are the basic intransitive, extended intransitive, transitive and ditransitive verbs, which are categorized by their valency. The remaining three verb classes are auxiliary verbs, complementizing verbs and compound verbs, the latter of which may be also subdivided into positional compound verbs and nominal compound verbs. Intransitive and transitive verbs constitute an open class, while the rest of the verb classes are closed.\footnote{N.b.: there needs to be more research conducted on the nature of the extended intransitive class, as I have not yet found out whether it is closed or open.} A list of all closed class verbs is presented in appendix C. An overview of verb classes is depicted in table \ref{t:verbclass}.

\begin{table}[H]
    \centering
    \begin{tabular}{@{}lll@{}}
    \toprule
    \multicolumn{1}{c}{Class} & \multicolumn{1}{c}{\begin{tabular}[c]{@{}c@{}}Number \\ of Arguments\end{tabular}} & \multicolumn{1}{c}{Argument Marking} \\ \midrule
    Intransitive              & 1                                                                                  & S-∅ V                                \\
    Ext. Intransitive         & 2                                                                                  & S-∅ V E-{\Dat} / {\Loc}                      \\
    Transitive                &                                                                                    &                                      \\
    \null\quad a) Core Transitives       & 2                                                                                  & A-{\Erg} V P-∅                          \\
    \null\quad b) Non-Core Transitives   & 2                                                                                  & A-{\Erg} V P-\textsc{ncore}                      \\
    Pos. Compound Verbs       & 1-2                                                                                & S-∅ V or A-{\Erg} V P-∅                 \\
    Ditransitive              & 3                                                                                  & D-{\Erg} T-∅ V R-{\Dat}                    \\ \bottomrule
    \end{tabular}
    \caption{Verb Classes Overview}
    \label{t:verbclass}
\end{table}
\paragraph{Intransitive Verbs}
Intransitive verbs are verbs which take one S argument, as can be seen in example (\ref{ex:intr1}). This argument, presupposing it is not a pronoun, is left unmarked, i.e. takes the absolutive null morpheme. If the argument is a pronoun, it must take its S form, as can be seen in example (\ref{ex:intr2}).
\begin{examples}
    \ex \label{ex:intr1}
    \words tahai sóíhikóa\MC3
    \bits tahai sóí -hi =kóa
    \gloss goat sleep RPASTT DECL
    \tr The goat slept
    \ex \label{ex:intr2}
    \words pói sóíhikóa\MC3
    \bits pói sóí -hi =kóa
    \gloss SS:SUBJECT sleep RPASTT DECL 
    \tr You slept
\end{examples}
\paragraph{Extended Intransitive Verbs}
Extended intransitive verbs are verbs which take two arguments, an S argument and an E argument. The S argument takes the absolutive case morpheme or the S pronoun form, whilst the E argument takes the dative case or the dative pronoun form. Some examples are shown in (\ref{ex:exitr1}) and (\ref{ex:exitr2}). Some extended intransitive verbs related to movement, such as \emh{háa} `to go' or \emh{póói} `to come' take the locative case for their E argument, as shown in (\ref{ex:exitr3}). There are \textit{5} extended intransitive verbs in the analyzed corpus. Some explanations concerning (\ref{ex:exitr2}) and (\ref{ex:exitr3}): in the former, the speaker tells us about a ritual in which a mother is applying a special facial paint to her son; in the latter, the speaker is talking to his fellow huntsman, mentioning how one of their dogs was playfully following a bird.
\begin{examples}
    \ex \label{ex:exitr1} 
    \words kahoa hókaáhàòkóa\MC3 íákìih\MC2
    \bits kahoa hókaá -hàò =kóa íákí -ìih 
    \gloss turtle live.in DPASTT DECL tree.hole DAT
    \tr The turtle lives in a tree hole
    \ex \label{ex:exitr2}
    \words xapisóoha\MC2 hóaxáko sahóóhaká\MC3 kòh
    \bits xapisóo =à hóaxáko sahóó -a =ká kòh
    \gloss mother LINK TT apply.color ST DECL TSM:DAT
    \tr His mother was painting his [face]
    \ex \label{ex:exitr3}
    \words tài háahaka\MC3 paátihóo\MC2
    \bits  tài háa -a =ka paáti -hóo 
    \gloss DEM:PROX go ST DECL type.of.bird LOC
    \tr This one is following a macaw
\end{examples}
\paragraph{Transitive and Ditransitive Verbs}\label{s:trans}
Transitive verbs are verbs which take two arguments, an A and an P argument. If the arguments of a transitive verb are pronominal, they take their respective A and P form. A noun in the A function takes the ergative case, marked by the morpheme \emh{-(V́)h}, while a noun in the P function takes the unmarked absolutive case. Examples of the usage of transitive verbs and their arguments' marking can be seen in (\ref{ex:tr1}) and (\ref{ex:tr2}). In (\ref{ex:tr1}) the speaker recounts a story in which a negligent boy mislaid his ritual loincloth, only to be found by the speaker; the speaker is surprised as she sees mice chewing on the cloth.
\begin{examples}
    \ex \label{ex:tr1}
    \words hòotaháh\MC3 kíxohxíkóa\MC3 hása
    \bits hòotah -a -h kíxoh -xí =kóa hása 
    \gloss mouse PL ERG chew RPASTO DECL loincloth
    \tr The mice chewed the loincloth
    %\source \parencite[240]{changingvalency}
    \ex \label{ex:tr2}
    \words xah kíxohxíkóa\MC3 kòà
    \bits xah kíxoh -xí =kóa kòà
    \gloss TP:AGENT chew RPASTO DECL TSN:PATIENT
    \tr They chewed it
    %\source \parencite[240]{changingvalency}
\end{examples} 
However there are exceptions to this rule; some transitive verbs' objects do not take the absolutive case but a non-core or relational case instead. An example for such a verb is \emh{títo} `to listen, to smell', which takes an additional argument in the `next to' relational case, marked by the morpheme \emh{-(h)aí}. An example for the usage of those non-core transitive verbs is given in (\ref{ex:tr3}).
\begin{examples}
    \ex \label{ex:tr3}
    \words háá tóópahaí\MC2 títoi\MC2 tóóhaká\MC3
    \bits háá tóópa -haí títo -i tóó -a =ká 
    \gloss EPIST song NEXT listen DEP PROG ST DECL
    \tr Maybe she's listening to the song!
\end{examples}
Ditransitive verbs are verbs which take three arguments; those arguments are called D, T and R, and refer to the donor, theme and recipient of the verb. For nouns, the D function is marked by the ergative case, the T function is marked by the absolutive case and the R function is marked by the dative case. For pronouns, D and T correspond to the A and P forms, while the dative forms are used for the R function. There are \textit{2} ditransitive verbs in the analyzed corpus. Examples (\ref{ex:ditr1}) and (\ref{ex:ditr2}) showcase the usage of ditransitive verbs. 
\begin{examples}
    \ex \label{ex:ditr1}
    \words tííkóh\MC2 kóóh hóíhikóa\MC3 tatìih\MC2
    \bits tííko -h kóóh hóí -hi =kóa tati -h
    \gloss man's.name ERG fruit give RPASTT DECL child DAT
    \tr Diego gives a fruit to [his] child (for good)
    \ex \label{ex:ditr2}
    \words kó kòà hóíhikóa\MC3 hòah
    \bits kó kòà hóí -hi =kóa hòah
    \gloss TSM:AGENT TSN:PATIENT give RPASTT DECL TSF:DAT
    \tr He gives it to her
\end{examples}
The R of a ditansitive verb might also be marked by the `on top of' relational case; in that case, the action is temporary, while with the dative case, the action is permanent.\footnote{This obviously only applies if the verb can bear such a type of aspect; for example  with the verb \emh{taih}, `to thank sb. for sth.' this distinction wouldn't make sense.} This  usage is exemplfied in (\ref{ex:ditr3}). 
\begin{examples}
    \ex \label{ex:ditr3}
    \words tííkoh\MC2 kaxií hóíhikóa\MC3 tatìxa\MC2
    \bits tííko -h kaxií hóí -hi =kóa tati -xa
    \gloss man's.name ERG idol give RPASTT DECL child ONT
    \tr Diego gives the idol to [his] child (as a loan)
\end{examples}
\paragraph{Auxiliary Verbs}
Auxiliary verbs are verbs which mark aspect, negation and modality. there are two types of auxiliaries, non-affixing auxiliary verbs and affixing auxiliary verbs. I will first discuss the former, and subsequently move onto the latter. Non-affixing auxiliaries, despite their name, absorb all inflection from the lexical head. They encode negation and aspect. The lexical head then takes the dependent form, marked by the suffix \emh{-i}. There are \textit{3} non-affixing auxiliaries in the analyzed corpus of \textit{5} auxiliaries. (\ref{ex:aux1}) exemplifies the usage of non-affixing auxiliaries.
\begin{examples}
    \ex \label{ex:aux1}
    \words hóh\MC2 kìkahoó\MC2 hóikai\MC2 tóóxíkóa\MC4 hákiihahóo\MC4
    \bits hó -h kì- kahoó hóika -i tóó -xí -∅ =kóa h- ákiih -a -hóo
    \gloss man's.name ERG TSPOSS boat build DEP AUX:PROG RPASTO ST DECL FSPOSS see NMZ LOC
    \tr Joe was constructing his boat, when I had seen him.
    %\source \parencite[18]{smoyd1308}
\end{examples}
Affixing auxiliary verbs use an affixing process, in which the lexical head is attached to the auxiliary. The inflection of the verb is absorbed by the general auxiliary \emh{tàa}. There are \textit{1} affixing auxiliaries in the analyzed corpus of \textit{5} auxiliaries. An example of their usage can be seen in example (\ref{ex:aux2}).
\begin{examples}
    \ex \label{ex:aux2}
    \words páó híí kaxáhákoo\MC2 tàahaka\MC3
    \bits páó híí kaxá= hákoo tàa -a =ka 
    \gloss SS:AGENT meal eat OBLIGATIVE AUX ST DECL
    \tr You should eat your meal (or else you will be hungry later)
\end{examples}
To negate an auxiliary verb, instead of using the negation auxiliary, the disjunct prefix \emh{kaí=}, which is derived from the past negative auxiliary \emh{kaíhao}, is used. Example (\ref{ex:aux3}) showcases this negation strategy.
\begin{examples}
    \ex \label{ex:aux3}
    \words kóíhihóo\MC2 ahá háakaíhákoo\MC3 atàa\MC2
    \bits kóíhi -hóo ahá háa= kaí= hákoo a- tàa 
    \gloss forest LOC alone go NEG OBLIGATIVE PASS AUX
    \tr One shouldn't go into the forest alone.
\end{examples}
\paragraph{Complementizer Verbs}
Complementizer verbs are verbs which take an entire clause as their object, such as e.g. \emh{áa} `to say (that)' or \emh{óa} `to think'. They may take all normal inflectional affixes. To mark the end of a complement clause, the complementizers \emh{kóó-} or \emh{kóhii} are used.\footnote{More on the usage of complementizers  in section \ref{s:complclaus}}. The usage of complementizer verbs is exemplified in (\ref{ex:complv1}).
\begin{examples}
    \ex \label{ex:complv1}
    \words kóíhihóo\MC2 ahá háakaíhákoo\MC3 atàa\MC2 kóhii káixoáh\MC2 áatóhikóa\MC3
    \bits kóíhi -hóo ahá háa= kaí= hákoo a- tàa kóhii káixo -áh áa -tóhi =kóa
    \gloss forest LOC alone go NEG OBLIGATIVE PASS AUX C mother AUG say DPASTO DECL
    \tr [My] grandmother said that one shouldn't go into the forest alone
\end{examples}
While the complementizers mark indirect speech, there is a quotative marker \emh{ko} to mark direct speech. An example of this technique can be found in (\ref{ex:complv2}).
\begin{examples}
    \ex \label{ex:complv2}
    \words `kóíhihóo\MC2 ahá háakaíhákoo\MC3 atàa'\MC2 ko káixoáh\MC2 áatóhikóa\MC3
    \bits kóíhi -hóo ahá háa= kaí= hákoo a- tàa ko  káixo -áh áa -tóhi =kóa
    \gloss forest LOC alone go NEG OBLIGATIVE PASS AUX QUOTATIVE mother AUG say DPASTO DECL
    \tr [My] grandmother said: ``One shouldn't go into the forest alone''
\end{examples}
\paragraph{Positional Compound Verbs}
Positional compound verbs are verbs with a bipartite stem. The first part (called V1) is from a relatively open class of ca. 100 members, while the second part (called DTR) is from a closed class of 10 members. They inflect like normal intransitive or transitive verbs. The DTRs depict explicit or implied motion in a certain \textbf{D}irection, into a \textbf{T}opological \textbf{R}elation. There are four subclasses of V1s, classificatory (type of figure or ground), dispositional (posture or position of the figure), means (manner or means of change), and trajectory (shape of path of change). While the first of these constitutes an open class, the rest of them are closed class. A full list of all V1s and DTRs can be found in appendix D. Examples (\ref{ex:vposit1}) and (\ref{ex:vposit2}) showcase the usage of transitive and intransitive positional compound verbs.
\begin{examples}
\newbaarucmd{\cl}{\baarujuncture{\texttt{==}}}
    \ex \label{ex:vposit1}
    \words hoa tapáxi xóhiitohíxíkóato\MC5 hííkohóo\MC2
    \bits hoa tapáxi xóhii- tohí -xí =kóa \cl to hííkoh -hóo
    \gloss TSF:AGENT honey liquid in RPASTO DECL VIS bowl LOC
    \tr She poured the honey into the bowl (I saw it)
    \ex \label{ex:vposit2}
    \words (íí) saháxìhikóa\MC4
    \bits (íí) sahá- xì -hi =kóa 
    \gloss (TSN:SUBJECT) fall away RPASTT DECL
    \tr It (the leaf) fell away from me, i.e. it was blown away by the wind
\end{examples}

\subsection{Mode}\label{svmode}
In the following section I will present the mode slot within the verbal stem. Mode is one of many verbal grammatical categories, and it marks some modal expression, most importantly, imperatives, optatives and hortatives. Interrogative semantics are also covered by the mode prefixes. 
\subsubsection{Interrogative \emh{í-}}
The interrogative mode is used to mark questions, whether it be polar or content questions. It is marked by the prefix \emh{í-}. If the verb is in the interrogative mode, a special set of person markers are used (see section \ref{svperson}). Some examples of the interrogative mode's employment can be seen in (\ref{ex:intrg1}) and (\ref{ex:intrg2}). More information about interrogative clauses, especially in a syntactic context, is given in chapter \ref{c:interrogative}.
\begin{examples}
    \ex \label{ex:intrg1}
    \words {kóha haípi} há tááhi\MC2 ítóóhah\MC3
    \bits {kóha haípi} há tááh -i í- tóó -ah
    \gloss tea EGO drink DEP INTRG AUX ST 
    \tr Are you drinking tea?
    \ex \label{ex:intrg2}
    \words páhisòì\MC2 há íxóatiah\MC3 
    \bits páhi -sòì há í- xóati -ah
    \gloss what QU EGO INTRG search.for ST
    \tr What are you searching for?
\end{examples}
\subsubsection{Imperatives}
There are three imperative prefixes in the Hapi language, all of which occupy the mode slot. There are two positive markers, distal and proximal, and one deictic neutral negative marker. The distal and proximal prefixes are \emh{tà-} and \emh{hi-}. The distal morphemes are used when the command has a special distance in space — 'do there' —, while the proximal morphemes are used for contexts where the command's execution is imminent. This does not only apply for spatial, but also for temporal deixis; the distinction is exemplified in (\ref{ex:impv1}) and (\ref{ex:impv2}). 
\begin{examples}
    \ex \label{ex:impv1}
    \words tàhokó hikakía\MC3
    \bits tàhokó  hi- kakí -a 
    \gloss FAT:DAT DIST.POSIT.IMPV wait ST
    \tr You, wait for us (in some distant time or place)
    %\source \parencite[186a]{aikhenamazon}
    \ex \label{ex:impv2}
    \words tàhokó tàkakía\MC3
    \bits tàhokó  tà- kakí -a 
    \gloss FAT:DAT PROX.POSIT.IMPV wait ST
    \tr You, wait for us (here and now)
    %\source \parencite[186b]{aikhenamazon}
\end{examples}
To exclude either the spatial or temporal sense from the distal imperative, spatial or temporal adverbs can be employed, as showcased in (\ref{ex:impv3}) and (\ref{ex:impv4}).
\begin{examples}
    \ex \label{ex:impv3}
    \words tàhokó hih hikakía\MC3
    \bits tàhokó hih hi- kakí -a
    \gloss FAT:DAT there.NVIS DIST.POSIT.IMPV wait ST
    \tr You, wait for us (in some distant place, but right now)
    \ex \label{ex:impv4}
    \words tàhokó sóso hikakía\MC3
    \bits tàhokó sóso hi- kakí -a
    \gloss FAT:DAT then DIST.POSIT.IMPV wait ST
    \tr You, wait for us (in some distant time, but here)
\end{examples}
The negative imperative \emh{xi-} is used for negated commands. Examples (\ref{ex:impv5}) present this process.
\begin{examples}
    \ex \label{ex:impv5}
    \words tàhokó xikakía\MC3
    \bits tàhokó xi- kakí -a
    \gloss FAT:DAT NEGAT.IMPV wait ST
    \tr Do not wait for us!
\end{examples}
\subsubsection{Jussive \emh{háá-}}
The jussive mode, marked by the prefix \emh{háá-} marks a variety of semantics on the verb. Verbs that are in the jussive mode do not take any other markers besides the jussive, neither person nor tense morphemes. Firstly, it may be used as a third-person imperative, having the speaker convey appeal to the adressee to enable an event to occur; this is exemplified in (\ref{ex:juss1}) and (\ref{ex:juss2}). The speaker expresses a desire that the event will take place, but unlike with the imperative, the speaker has limited control over the event's outcome. In this case, the affected third-person participant takes the semantic role of an agent or a subject, depending on the transitivity of the verb. (\ref{ex:juss2}) was uttered in the context of a woman who wanted her husband to come help in the garden, although he has been up all night drinking.
\begin{examples}
    \ex \label{ex:juss1}
    \words xipásaa\MC3 xixíxí\MC2 kòà kó hááxíxí\MC2
    \bits xi- pása -a xi- xíxí kòà kó háá- xíxí
    \gloss NEGAT.IMPV carry ST NEGAT.IMPV go.out TSM:PATIENT TSM:SUBJECT JUSS go.out
    \tr Don't carry him out (of the house), let him go out by himself
    %\source \parencite[339a]{hupepps}
    \ex \label{ex:juss2}
    \words tá kó háásóí\MC2
    \bits tá kó háá- sóí
    \gloss later TSM:SUBJECT JUSS sleep 
    \tr Let him sleep later
    %\source \parencite[339b]{hupepps}
\end{examples}
Another function of the jussive prefix is comparable to an optative in other languages. The speaker again expresses a desire for the event to happen, but in the optative case the speaker has little to no control over its outcome. An example is given in (\ref{ex:juss3}) and (\ref{ex:juss4})
\begin{examples}
    \ex \label{ex:juss3}
    \words pói hááhooxá\MC2
    \bits pói háá- hooxá 
    \gloss SS:SUBJECT JUSS be.glad
    \tr May you live merrily!
    \ex \label{ex:juss4}
    \words pípíò\MC2 hááxàa\MC2
    \bits pípí -ò háá- xàa
    \gloss rich CL:anim.sg JUSS be 
    \tr If only I were rich!
\end{examples}
Another usage of the optative function is in complement clauses, when the verb which initializes the clause possesses semantics of wishing and hoping. See (\ref{ex:juss5}) for an example.
\begin{examples}
    \ex \label{ex:juss5}
    \words káakaahikóa\MC3 kóo\MC2 hááháahi\MC3 \\ hááxíihaxi\MC2 háátíto\MC2
    \bits káakaa -hi =kóa kóó -∅ háá- áa -hi \\ háá- xíihaxi háá- títo
    \gloss hope RPASTT DECL C FF JUSS say RPASTT JUSS tell.a.story JUSS listen
    \tr I myself was hoping that you would tell me a story (before dinner)
\end{examples}
Lastly, the jussive mode may express a speaker'ss disengagement from a negatively viewed event, in which case the speaker voices a feigned indifference, associated with a lack of control, as exemplified in (\ref{ex:juss6}) and (\ref{ex:juss7}).
\begin{examples}
    \ex \label{ex:juss6}
    \words kó háákatí\MC2
    \bits kó háá- katí
    \gloss TSM:SUBJECT JUSS dance
    \tr Let him dance! (uttered by a jealous wife) 
    %\source \parencite[340a]{hupepps}
    \ex \label{ex:juss7}
    \words íí háákíío\MC2 tàhopí xàa koítaó\MC2
    \bits íí háá- kíío tàhopí xàa koí -taó 
    \gloss TSN:SUBJECT JUSS rain FAS COP hut BEL
    \tr Let it rain, we're in (lit.: below) a house
    %\source \parencite[340b]{hupepps}
\end{examples}
\subsection{Valency}\label{svvalency}
In the following chapters I will discuss the valency-modifying morphemes which are marked directly on the Hapi verb. This includes the antipassive section (\ref{s:antip}), the passive section (\ref{s:pass}) and the causative section (\ref{s:caus}) prefixes. 
\subsubsection{Antipassive \emh{h(V)-}}\label{s:antip}
The Hapi antipassive, marked by the prefix \emh{h(V)-}, is used to turn a transitive verb into an intransitive one. This is achieved by demoting the former A of the transitive verb to S and omitting the O argument. A straightforward transitive clause can be seen in example (\ref{ex:antip1}).
\begin{examples}
    \ex \label{ex:antip1}
    \words hápaáh xósóóhaká\MC3 hapi
    \bits hápaáh xósóó -a =ká hapi
    \gloss dog:ERG bite ST DECL people
    \tr The dog bites people
    %\source \parencite[232a]{aikhenamazon}
\end{examples}
If the prefix \emh{h(V)-} is added, the verb becomes intransitive, the underlying A turns into an S and the O is omitted. In (\ref{ex:antip2}), 'biting' is what the dog does, no matter whom it may bite.
\begin{examples} 
    \ex \label{ex:antip2}
    \words hápaáh hóxósóóhaká\MC4
    \bits hápaáh hó- xósóó -a =ká 
    \gloss dog:ERG ANTIP bite ST DECL
    \tr The dog bites
    %\source \parencite[232b]{aikhenamazon}
\end{examples}
However, there's a second side to the Hapi antipassive. When applied to intransitive verbs, it imparts a habitual, 'characteristic' meaning, as can be inferred from example (\ref{ex:antip3}).
\begin{examples}
    \ex \label{ex:antip3}
    \words (íí) hípíhaahaká\MC4
    \bits (íí) hí- píhaa -a =ká
    \gloss TSN:SUBJECT ANTIP jump ST DECL
    \tr It jumps (speaking of a species of titi monkeys)
\end{examples}
\subsubsection{Passive \emh{a-}}\label{s:pass}
Just like the antipassive, Hapi's passive derives an intransitive clause from a transitive one, however instead of omitting the underlying O, it demotes it to the S of the surface clause. The former A argument hereby goes into a peripheral function and may be marked by the perlative case.\footnote{Cf. section \ref{s:perlative}} In Hapi, the passive is marked by the prefix \emh{a-}. In (\ref{ex:passive1}) we have a straightforward transitive sentence, whose passive variant is showcased in example (\ref{ex:passive2}). Note that the underlying A needn't be mentioned at all, and can be omitted, as shown in example (\ref{ex:passive3})
\begin{examples}
    \ex \label{ex:passive1}
    \words soátáh\MC2 kíípáatóhikóa\MC3 kaókáoxii\MC2
    \bits soáta -h kíípáa -tóhi =kóa kaókáo -xii
    \gloss type.of.possum ERG kill DPASTO DECL duck DIM
    \tr A possum killed a duckling
    \ex \label{ex:passive2}
    \words kaókáoxii\MC2 akíípáatóhikóa\MC4 soátatah\MC2
    \bits kaókáo -xii a- kíípáa -tóhi =kóa soáta -tah
    \gloss duck DIM PASS kill DPASTO DECL type.of.possum PRL
    \tr The duckling was killed by a possum
    \ex \label{ex:passive3}
    \words kaókáoxii\MC2 akíípáatóhikóa\MC4
    \bits kaókáo -xii a- kíípáa -tóhi =kóa
    \gloss duck DIM PASS kill DPASTO DECL
    \tr The duckling was killed
\end{examples}
However \emh{a-} has other usages than the prototypical passive as well. In (\ref{ex:passive4}) the reflexive meaning is showcased. 
\begin{examples}
    \ex \label{ex:passive4}
    \words asaxíká\MC4 sáha sóh
    \bits a- sa -xí =ká sáha sóh 
    \gloss REFL cut RPASTO DECL knife with
    \tr (I) cut myself with a knife
    %\source \parencite[229a]{aikhenamazon}
\end{examples}
Another function of this morpheme is, similar to the reflexive, reciprocal semantics. If the subject is plural, it is likely to mark a reciprocal, as can be seen in (\ref{ex:passive5}).
\begin{examples}
    \ex \label{ex:passive5}
    \words xah asaxíká\MC4 sáha sóh
    \bits xah a- sa -xí =ká sáha sóh 
    \gloss TP:SUBJECT RECIP cut RPASTO DECL knife with
    \tr They cut each other with a knife
    %\source \parencite[229b]{aikhenamazon}
\end{examples}
Nevertheless, it isn't always clear whether a reflexive or reciprocal meaning is implied; in that case, the context and pragmatics give further clues. In example (\ref{ex:passive6}), the speaker tells us about some children, playing a game in which they need to group in pairs and the children need to choose each other. The reflexive meaning is not appropriate: the children did nothing to themselves.
\begin{examples}
    \ex \label{ex:passive6}
    \words haí xah asihoaka\MC4
    \bits haí xah a- siho -a =ka 
    \gloss so TP:SUBJECT RECIP choose ST DECL
    \tr So they are choosing each other
    \tr:* So they are choosing themselves
\end{examples} 
\subsubsection{Causative \emh{-áh} and \emh{-kó}}\label{s:caus}
There are two types of causative constructions in the Hapi language; one using the suffix \emh{-áh} to derive transitives from intransitives and another one, \emh{-kó}, which forms ditransitives from transitives. I will firstly discuss the former. Here, the argument in underlying S function (`the causee') goes into O function in the causative, and a new argument (`the causer') is introduced in A function. In (\ref{ex:caus1}), a simple intransitive clause is shown, and in example (\ref{ex:caus2}) the correspondent transitive clause is presented.
\begin{examples}
    \ex \label{ex:caus1}
    \words sóíhakaxa\MC4
    \bits sóí -a =ka =xa
    \gloss sleep ST DECL CONT
    \tr (He) is sleeping
    %\source \parencite[237a]{aikhenamazon}
    \ex \label{ex:caus2}
    \words óxííₒ sóíhaáhkaxa\MC5
    \bits óxííₒ sóí -a -áh =ka =xa
    \gloss child sleep ST CAUS:ITR DECL CONT 
    \tr (He) is putting the child to sleep.
    %\source \parencite[237b]{aikhenamazon}
\end{examples}
When deriving a ditransitive clause from a transitive one, the suffix \emh{-kó} is used. (\ref{ex:caus3}) and (\ref{ex:caus4}) exemplify this.
\begin{examples}
    \ex \label{ex:caus3}
    \words pápo hóhixíkóa\MC3 kó
    \bits pápo hóhi -xí =kóa kó
    \gloss paper take RPASTO DECL TSM:AGENT
    \tr He took the paper
    %\source \parencite[237c]{aikhenamazon}
    \ex \label{ex:caus4}
    \words sóso pápo kòà hóhixíkókóatì\MC5 kó
    \bits sóso pápoₒ kòàₒ hóhi -xí -kó =kóa =tì kóₐ 
    \gloss then paper TSM:PATIENT take RPASTO CAUS:TRANS DECL \textsc{\textup{again}} TSM:AGENT
    \tr Then he sent the paper back to him again
    \tr Then he caused him to take the paper back again
    %\source \parencite[237d]{aikhenamazon}
\end{examples}
A 'sociative' causative, marked by the causative and the comitative coordinator \emh{sóh} in combination with a personal pronoun, may have overtones of the causer's involvement; compare the straightforward example (\ref{ex:caus5}) with the sociative construction in (\ref{ex:caus6}).
\begin{examples}
    \ex \label{ex:caus5}
    \words kìhaóxí\MC2 haíhíkihikókóa\MC5 pàahtah\MC2
    \bits kì- haóxí haí- híki -hi -kó =kóa pàah -tah
    \gloss TSPOSS son CL:anim.sg across RPASTT CAUS:TRANS DECL river PRL 
    \tr He caused his son to cross the river \\ (ordered him without being involved)
    %\source \parencite[240a]{aikhenamazon}
    \ex \label{ex:caus6}
    \words kó sóh [kìhaóxí]ₒ\MC2 haíhíki-\MC2 \\ hikókóa\MC3 [pàahtah]ₒ\MC2
    \bits kó sóh kì- haóxí haí- híki hi -kó =kóa pàah -tah 
    \gloss TSM with TSPOSS son CL:anim.sg across RPASTT CAUS:TRANS DECL river PRL
    \tr He caused his son to cross the river (helped him across the river)
    %\source \parencite[240b]{aikhenamazon}
\end{examples}
\subsection{Past tenses}\label{svpast}
In the following section I will consider Hapi's elaborate system of past tenses. First, I will present the recent past ({\Rpasto}), then I will move on and discuss the three other past tenses, namely, the intermediate past ({\Rpastt}), the distant past ({\Dpasto}) and finally the remote past ({\Dpastt}). An overview of the past tenses and there deictic scopes can be found in table \ref{t:pasttenses}.
\begin{table}[H]
    \centering
    \begin{tabular}{@{}lllll@{}}
    \toprule
    Section  & Form  & Gloss    & \begin{tabular}[c]{@{}l@{}}Minimum\\ distance\end{tabular}                  & \begin{tabular}[c]{@{}l@{}}Maximum\\ distance\end{tabular} \\ \midrule
    \ref{s:recpast} & -xí   & {\Rpasto}  & \makecell[lt]{immediately prior \\ to speech time} & three days ago                                                \\
    \ref{s:intpast} & -hi   & {\Rpastt}  & yesterday                                                                   & last week                                                  \\
    \ref{s:distpast} & -tóhi & {\Dpasto} & last week                                                                   & years ago                                                  \\
    \ref{s:rempast} & -hàò  & {\Dpastt}  & \multicolumn{2}{l}{historical events and narratives}                                                                                     \\ \bottomrule
    \end{tabular}
    \caption{Past Tenses Overview}
    \label{t:pasttenses}
  \end{table}
\subsubsection{Recent Past \emh{-xí}}\label{s:recpast}
The recent past, marked by the morpheme \emh{-xí}, encodes that an event took place immediately prior and up to three days prior to the speech time. An absolute reading yields the encoding of an event that took place two days prior to the speech act. In example (\ref{ex:recpast1}), the primary function of the recent past marker is showcased. In the following narrative, the speaker talks about going on a fishing trip with his friend Atah.
\begin{examples}
    \ex \label{ex:recpast1}
    \words hitaóha tà atah sóh háaha\MC2 ahihxíkóa\MC3
    \bits hitaóha tà atah sóh háa -ha ahih -xí =kóa
    \gloss first FS:SUBJECT man's.name with go NMZ think.about RPASTO DECL
    \tr First I thought about going out with Atah
\end{examples}
In discourse, the recent past marker is repeated throughout the narrative, as long as events occur within the temporal deictic range of the marker and are described in-sequence.\footnote{Then, the sequential marker \emh{==ó} may be used to link the predicates.} (\ref{ex:recpast2}) and (\ref{ex:recpast3}) are the continuation of (\ref{ex:recpast1}), and they show that \emh{-xí} maintains its position within the predicate and displays discourse perseverance.
\begin{examples}
    \ex \label{ex:recpast2}
    \words haí kaxáxíkoi\MC3 atah sóh tà xíxíxí\MC2 kókihíkixí\MC2 kàahxí\MC2 koíháaxíkóa\MC3
    \bits haí kaxá -xí =koi atah sóh tà xíxí -xí kókihíki -xí kàah -xí koíháa -xí =kóa  
    \gloss then eat RPASTO \textsc{\textup{S/A>S(PE)}} man's.name with FS:SUBJECT go.away RPASTO fish RPASTO kill RPASTO go.home RPASTO DECL
    \tr After eating, I went with Atah to go fishing
    \ex \label{ex:recpast3}
    \words hitaóha kaóhtiihóo\MC2 tàhokó kòhóohixíkóa\MC3
    \bits hitaóha kaóhtii -hóo tàhokó kòhóohi -xí =kóa 
    \gloss first uncle LOC FAT:SUBJECT arrive RPASTO DECL
    \tr Firstly, we arrived at my uncle's
\end{examples}
The recent past marker may also be used in a relative temporal sense, i.e. relative to the framing narrative. This can be seen in example (\ref{ex:recpast4}). Here, the framing narrative is the event of the parents speaking, which itself is set in the distant past tense. However, the event which is described in the speech act, namely, the coming of age of their children, is put in the recent past, even though in an absolute sense it has happened years ago.
\begin{examples}
    \ex \label{ex:recpast4}
    \words `tótihaí\MC2 kìotixíkóa'\MC3 ko \\ kìkáota\MC2 sóh kìxapisóo\MC2 áatóhikóa\MC3
    \bits `tóti -haí kìoti -xí =kóa' ko kì- káota sóh kì- xapisóo áa -tóhi =kóa
    \gloss prepared CL:anim.pl become RPASTO DECL QUOTATIVE TSPOSS father and TSPOSS mother say DPASTO DECL
    \tr Her father and mother said: ``They (their children) have (already) become adolescents''
\end{examples}
\subsubsection{Intermediate Past \emh{-hi}}\label{s:intpast}
The intermediate past conjunct suffix \emh{-hi} encodes that an event took place between yesterday (minimum) and one week (maximum) from the time of speech or some reference time. In (\ref{ex:intpast1}), the primary function of the intermediate past is exemplified. In the following, the speaker tells us about his pet armadillo, which escaped from the speaker's hut. 
\begin{examples}
    \ex \label{ex:intpast1}
    \words hákiikíípoh\MC3 saahì\MC2 tóhì\MC2 kohsoa\MC2 hakó tatáakai xíxíhi\MC2 kohápahikoa\MC3
    \bits há- kiikíí =poh saa -hì tóh -hì koh -soa hakó tataákai xíxí -hi kohápa -hi =kóa
    \gloss FSPOSS armadillo TOPIC two CL:time three CL:time day PL ago do.suddenly.ITR go.away RPASTT leave RPASTT DECL
    \tr As for my (pet) armadillo, two or three days ago, it suddenly disappeared (lit. went away and left)
\end{examples}
In narratives set in the intermediate past, \emh{-hi} exhibits discourse perseverance, if and only if the narrated events fall into the scope of the marker, i.e. occur between yesterday and one week prior to the speech act. This can be seen in the examples (\ref{ex:intpast2}) and (\ref{ex:intpast3}), which are a continuation of the narrative introduced in (\ref{ex:intpast1}).
\begin{examples}
    \ex \label{ex:intpast2}
    \words kìsaháxìhahóo\MC5 tàah ákiihikóahikí\MC4
    \bits kì- sahá -xì -ha -hóo tàah ákiih -hi =kóa =hikí 
    \gloss TSPOSS fall out.of NMZ LOC FS:AGENT see RPASTT DECL NOCT
    \tr (And) at night, when it (the armadillo) had already left (lit. fallen out), I didn't see (it anymore).
    \ex \label{ex:intpast3}
    \newbaarucmd{\cl}{\baarujuncture{\texttt{==}}}
    \words xóíi kipá sóh tàhokó \\ hóxóatihikóahó\MC5 tapíh\MC2 oháhiáhkóa\MC4 tohípisáahió\MC3 híó káhxìhikóakah\MC4
    \bits xóíi kipá sóh tàhokó hó- xóati -hi =kóa \cl ó tapí -h ohá -hi -áh =kóa tohípi -sáahi \cl ó híó káhxì -hi =kóa =kah
    \gloss FRUST:ITR woman's.name with FAT:SUBJECT ANTIP search RPASTT DECL SEQ man's.name ERG run RPASTT CAUS:ITR DECL willingness PVT SEQ but.DS catch RPASTT DECL PUNCT
    \tr I went in vain to search (it) with Kipá and Tapí let (it) escape but I caught it
\end{examples}
The intermediate past may also be employed in a relative temporal sense, i.e. when used in a framing narrative. This most often appears in speech reports, whether it be direct or indirect. An example for the relative usage of the intermediate past can be found in (\ref{ex:intpast4}). In this case, the framing narrative is set in the remote past, while the event that is described in the narrative occurs a few days prior; this means, that even though in absolute terms, the event occurs several years ago, it is still marked by the intermediate past.
\begin{examples}
    \ex \label{ex:intpast4}
    \newbaarucmd{\cl}{\baarujuncture{\texttt{==}}}
    \words áahàòkóa\MC3 kóó sóso tàiho \\ hahápia\MC3 sòihikóasah\MC4
    \bits áa -hàò =kóa kóó sóso tàiho ha- hápi -a sòi -hi =kóa \cl sah 
    \gloss say DPASTT DECL C then here ANTIP be.accustomed NMZ want RPASTT DECL INFER
    \tr A long time ago they said that apparently (they) wanted to live (with us) here (a couple of days ago)
\end{examples} 
\subsubsection{Distant Past \emh{-tóhi}}\label{s:distpast}
The distant past \emh{-tóhi} marks events that took place from last week up to years prior to the speech time. Contrary to the first set of past tense markers, i.e. the recent and intermediate past, the distant past (and additionally the remote past as well) cannot be used in a relative sense. An example concerning the primary usage of the distant past marker can be found in (\ref{ex:distpast1}).
\begin{examples}
    \ex \label{ex:distpast1}
    \words ahítáh kìtati\MC2 aháa ópoáhtóhikóa\MC3 
    \bits ahítáh kì- tati aháa ópoáh -tóhi =kóa 
    \gloss woman's.name:ERG TSPOSS child REP give.birth DPASTO DECL 
    \tr Ahíta gave birth to her child, it is said
\end{examples}
Once again, in narratives set in the distant past, discourse perseverance can be observed. In the continuation of (\ref{ex:distpast1}), (\ref{ex:distpast2}) and (\ref{ex:distpast3}) respectfully, the distant past marker appears yet again. 
\begin{examples}
    \ex \label{ex:distpast2}
    \words kìpóóihahóo\MC4 hoa \\ kaáxìtóhikóatíó\MC5 táakìxa\MC2
    \bits kì- póói -ha -hóo hoa kaá -xì -tóhi =kóa =tíó táaki -̀xa
    \gloss TSPOSS come NMZ LOC TSF:AGENT stand out.of DPASTO DECL INCH jar ONT
    \tr When her child was coming (out), she began putting into a jar
    \ex \label{ex:distpast3}
    \words táakihóo\MC2 kìkaáxìaxapa\MC5 xoátia\MC2 hoa toihtóhikóa\MC3
    \bits táaki -hóo kì- kaá -xì -a -xapa xoáti -a hoa toih -tóhi =kóa  
    \gloss jar LOC TSPOSS stand out.of NMZ PURP leg PL TSF:AGENT open DPASTO DECL
    \tr In order to put it in the jar, she spread (lit. opened) her legs (over the jar)
\end{examples}
\subsubsection{Remote Past \emh{-hàò}}\label{s:rempast}
The remote past \emh{-hàò} marks events that took place in mythological narratives, historical events and tales. As already mentioned, the remote past cannot me used as a relative tense. The primary usage of the remote past marker is showcased in (\ref{ex:rempast1}).
\begin{examples}
    \ex \label{ex:rempast1}
    \words kìákiihahóo\MC4 kó sóasihò\MC2 kìotihàòkóa\MC3 ò hìhkòoh hoa íhaíhàòkóa\MC3
    \bits kì- ákiih -a -hóo kó sóasih -ò kìoti -hàò =kóa ò hìhkòoh hoa íhaí -hàò =kóa 
    \gloss TSPOSS see NMZ LOC TSM:SUBJECT spirit CL:anim.sg transform DPASTT DECL and spouse.DAT TSF talk.to DPASTT DECL  
    \tr When (the sloth god) saw this, he transformed into a spirit and talked to the wife
\end{examples}
As with the other past tenses, discourse perseverance can be observed as long as the described events fit into the label of mythological narratives, historical happenings and tales. With (\ref{ex:rempast2}) and (\ref{ex:rempast3}) being the continuation of (\ref{ex:rempast1}), it can be seen that the marker \emh{-hàò} appears continuously.
\begin{examples}
    \ex \label{ex:rempast2}
    \words haí kó paixà hóhaó\MC2 kìsapííha\MC3 apákahàòkóa\MC3
    \bits haí kó paixà hóh -aó kì- sapíí -ha apáka -hàò =kóa
    \gloss then TSM:AGENT dirt penis BEL TSPOSS put NMZ propose DPASTT DECL
    \tr Then he (the sloth god) proposed (to her) to put dirt into (her husband's) penis
    \ex \label{ex:rempast3}
    \words hoa hih sóóhai\MC2 \\ hasapííhàòkóakohtóíh\MC5
    \bits hoa hih sóóha -i ha- sapíí -hàò =kóa =kohtóíh
    \gloss TSF:AGENT there:NVIS do.like.this ITR ANTIP put DPASTT DECL CRAST
    \tr The next night, she put (the dirt) there like this
\end{examples}
\subsection{Person agreement}\label{svperson}
In the following section I will present the verbal person agreement in the Hapi verb complex. Table \ref{t:vperson} presents an overview of the verbal agreement system. With the nominal case marking being ergative-absolutive, the pronoun system being split, the nominative-accusative personal agreement only adds to the complexity of Hapi's morphosyntactic alignment. 

\begin{table}[H]
    \centering
    \begin{tabular}{@{}lllll@{}}
    \toprule
        & \multicolumn{2}{l}{Non-Past}      & \multicolumn{2}{l}{Past}          \\ \midrule
        & Non-Interrogative & Interrogative & Non-Interrogative & Interrogative \\ \midrule
    1   & -h                & -∅            & -∅                & -a            \\
    2/3 & -a                & -ah           & -∅                & -a            \\ \bottomrule
    \end{tabular}
    \caption{Personal Agreement on Verbs}
    \label{t:vperson}
\end{table}

There are two general sets for verbal person markers, the non-past and past groups. While the non-past markers are used in present and future tenses, the past markers are used with the past tenses. Both of these groups are subdivided into non-interrogative and interrogative paradigms, the former of which being used in the declarative and imperative modes; as can be inferred from the table above, the second and third person markers are merged. Similarly, in the non-interrogative and past interrogative paradigms, all markers exhibit syncretism.
Examples (\ref{ex:vperson1}) - (\ref{ex:vperson3}) provide minimal examples for the non-interrogative paradigms.\footnote{I have excluded the plural pronouns in all of these examples for the sake of simplicity.}
\begin{examples}
    \ex \label{ex:vperson1}
    \words tà àihkóahóh\MC4
    \bits tà ài -h =kóa =óh
    \gloss FS:SUBJECT do FF DECL DIUR
    \tr During the day, I weave (lit. do)
    \ex \label{ex:vperson2}
    \words tà àihakóahóh\MC4
    \bits pói ài -a =kóa =óh
    \gloss FS:SUBJECT do FF DECL DIUR
    \tr During the day, you weave (lit. do)
    \ex \label{ex:vperson3}
    \words tà/pói/kó/hoa àixíkóahóh\MC5
    \bits tà/pói/kó/hoa ài -xí -∅ =kóa =óh
    \gloss FS{/}SS{/}TSM{/}TSF:SUBJECT do RPASTO FF{/}ST DECL DIUR
    \tr During the day, I/you/(s)he were/was weaving (lit. doing)
\end{examples}
Examples (\ref{ex:vperson4}) - (\ref{ex:vperson6}) provide similar examples for the interrogative paradigms.
\begin{examples}
    \newbaarucmd{\cl}{\baarujuncture{\texttt{==}}}
    \ex \label{ex:vperson4}
    \words páhiháaka\MC2 tà àikóa\MC3
    \bits páhi \cl áaka tà ài -∅ =kóa
    \gloss what MIR FS:SUBJECT do FF DECL
    \tr Of all things, what am I doing?
    \ex \label{ex:vperson5}
    \words páhiáaka\MC2 há pói àihahkóa\MC3
    \bits páhi \cl áaka há pói ài -ah =kóa 
    \gloss what MIR EGO SS:SUBJECT do ST DECL
    \tr Of all things, what are you doing?
    \ex \label{ex:vperson6}    
    \words páhiáaka\MC2 (há) tà/pói/kó/hoa àixíahkóa\MC4
    \bits páhi \cl áaka (há) tà/pói/kó/hoa ài -xí -ah =kóa 
    \gloss what MIR (EGO) FS{/}SS{/}TSM{/}TSF:SUBJECT do RPASTO FF{/}ST DECL
    \tr Of all things, what was/were I/you/(s)he doing?
\end{examples}
\begin{comment}
\subsection{Future tense}\label{svfut}
\section{Verb phrase structure}
\subsection{Order of verb phrase elements}\label{svorder}
\subsection{Second-position particles}\label{svparticles}
\subsection{Adverbs}\label{svadverbs}
\subsubsection{Lexical Adverbs}
\subsubsection{Verbal Adverbs}
\chapter{Adjectives and Adverbs}\label{c:adjectivesadverbs}
\section{Adjectives}
\subsection{Color Terms}
\subsection{Adnominal Adjectives}
\subsection{Adverbal Adjectives}
\section{Adverbs}
\subsection{Lexical Adverbs}
\subsubsection{Spatial Adverbs}
\subsubsection{Temporal Adverbs}
\subsubsection{Manner Adverbs}
\subsection{Verbal Adverbs}
\subsubsection{Overview}
\subsubsection{Transitive Class Adverbals}
\subsubsection{Intransitive Class Adverbals}
\subsubsection{Mixed Class Adverbals}
\chapter{Numerals}\label{c:numerals}
\section{Introduction}
\section{Cardinal Numerals}
\section{Ordinal Numerals}
\section{Collective Numerals}
\section{Multiplicative Numerals}
\section{Distributive Numerals}
\section{Other Quantifiers}
\subsection{PLACEHOLDER}
\chapter{Other Verbal Categories}
\section{Particles}
\subsection{PLACEHOLDER}
\section{Disjunct Affixes}
\subsection{PLACEHOLDER}
\section{Clitics}
\subsection{PLACEHOLDER}
\chapter{Derivation}
\section{Nominal}
\subsection{Noun-Noun Derivation}
\subsection{Others-Noun Derivation}
\section{Verbal}
\subsection{Verb-Verb Derivation}
\subsection{Others-Verb Derivation}
\section{Other Derivation}
\subsection{Adjectival Derivation}
\subsection{Adverbal Derivation}
\chapter{Clause-level Syntax}
\section{Constituent Order}
\section{Types of sentences}
\subsection{Declarative}
\subsubsection{Copular Clauses}
\subsubsection{Existential Clauses}
\subsubsection{Intransitive Clauses}
\subsubsection{Extended Intransitive Clauses}
\subsubsection{Transitive Clauses}
\subsubsection{Ditransitive Clauses}
\subsection{Interrogative}\label{c:interrogatives}
\subsubsection{Polar Questions}
\subsubsection{Question Words}
\subsection{Imperative}
\subsubsection{Imperative Clauses}
\subsubsection{Jussive Clauses}
\section{Multiverb Constructions}
\subsection{Serial Verb Constructions}
\subsection{Syndetic Verb Coordination}
\subsection{Switch-Reference Constructions}
\section{Subordination}
\subsection{Relative Clauses}
\subsection{Complement Clauses}\label{s:complclaus}
\subsection{Adverbal Clauses}
\subsubsection{Temporal/Conditional Clauses}
\subsubsection{Cause and Reason Clauses}
\subsubsection{Purposive Clauses}
\begin{appendices}
\chapter{List of inherently possessed Nouns}
\chapter{Classifiers}
\chapter{Closed class verbs}
\chapter{List of V1 and DTR components}
\end{appendices}
\end{comment}
%\printbibliography[title=Example Sentences Sources] 
% I commented out all the citations because I'm not sure yet whether I really want to have them in the document.
\end{document}

